---
title: "Figures and analysis for *Varroa Destructor* Inheritance Study"
author: "Nurit Eliash and Sasha Mikheyev"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    code_folding: hide
    theme: cerulean
    toc: yes
    toc_depth: 5
    toc_float:
      collapsed: yes
      df_print: paged
bibliography: ../references.bib
editor_options: 
  chunk_output_type: console
  markdown: 
    wrap: 72
---

```{=html}
<style type="text/css">
  body{
  font-size: 12pt;
}
</style>
```

```{r libraries, echo=TRUE, eval=TRUE, warning=FALSE, message=FALSE}
library("tidyverse")
library("dplyr")
library("ggplot2")
library("ggpubr")
library("scales")
library("ggpubr")
library("gridExtra")
library("grid")
library("GGally")
library("vcfR") # for extracting genotype data from a vcf file
library("janitor")
library("splitstackshape")
library("plotly")
library("kableExtra")
library("pastecs") # for stats stat.desc()
library("patchwork")
library("readr")  # for read_delim function
library(ggsignif) # for statistics of karyotyping

knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                     fig.width = 10,
                      fig.asp = 0.4,
                      out.width = "100%")

# setup-source-data
rmd_dir   <- dirname(rstudioapi::getActiveDocumentContext()$path)
source_dir <- file.path(rmd_dir, "Source_data")
if (!dir.exists(source_dir)) dir.create(source_dir, recursive = TRUE)
```

This document contains analyses and figures for our study on the mode of genetic inheritance in *Varroa destructor*.     
The full paper by Eliash et al. 2025, "Varroa mites escape the evolutionary trap of haplodiploidy", can be found in [here](https://www.biorxiv.org/content/10.1101/2024.05.10.593493v2).      

The document is composed of two main sections: 

- Modeling the Effective population size (Ne)      
- Genomic pedigree study   

# Modeling the Effective population size (Ne)

### Supplementary Figure 1: Model-predicted effective population size (Ne) 

This model starts each scenario from the same initial condition: a single
founding female mite (`N[1] = 1`). Population size then changes across
generations according to the same demographic rule, while only the
scenario parameters (`C` and `K`) are changed. Here, `C` is the carrying
capacity (the maximum sustainable number of reproducing females), and
`K` is the number of available brood cells. The dashed vertical line in
each panel marks the first generation at which the modeled population
reaches carrying capacity (`N = C`). Because each panel has its own `C`
and `K` values, this saturation point can occur at different generations.
Importantly, even when two scenarios have the same ratio (`C/K`), the
curves are not expected to be identical: the model depends on the
absolute values of `C` and `K` (not only their ratio), so changing scale
can change both population dynamics and the resulting effective population
size (`Ne`).

```{r model, echo=TRUE, eval=TRUE, warning=FALSE, message=FALSE}
# Revised model (April 2026) wrapped as a reusable function
run_varroa_model <- function(C, K, n = 2, p0 = 0.1, f1 = 0.1, Tt = 50) {
  # Demographic dynamics
  N <- rep(0, Tt)
  N[1] <- 1
  for (t in 2:(Tt - 1)) {
    N[t] <- round((1 + (n - 1) * (1 - N[t - 1] / C)) * N[t - 1])
  }
  P <- c(0, K / N * (1 - (1 - 1 / K)^N))

  # Cell occupancy probabilities
  P1 <- rep(0, Tt)
  for (i in 3:(Tt + 1)) {
    P1[i - 1] <- 1 / N[i - 2] / n * sum(sapply(
      1:N[i - 2],
      function(x) 1 / x^2 * choose(N[i - 2] - 1, x - 1) / K^(x - 1) * (1 - 1 / K)^(N[i - 2] - x)
    ))
  }
  P2 <- rep(0, Tt)
  for (i in 3:(Tt + 1)) {
    P2[i - 1] <- 1 / N[i - 2] * sum(sapply(
      1:N[i - 2],
      function(x) 1 / x * (1 - 1 / x / n) * choose(N[i - 2] - 1, x - 1) / K^(x - 1) * (1 - 1 / K)^(N[i - 2] - x)
    ))
  }
  P3 <- rep(0, Tt)
  for (i in 3:(Tt + 1)) {
    P3[i - 1] <- 1 / N[i - 2] * sum(sapply(
      2:N[i - 2],
      function(x) (1 - 1 / x) / x / n * choose(N[i - 2] - 1, x - 1) / K^(x - 1) * (1 - 1 / K)^(N[i - 2] - x)
    ))
  }
  P4 <- 1 - P1 - P2 - P3

  # -----------------------------
  # Model 1: haplo-diploidy
  # -----------------------------
  p1 <- rep(0, Tt)
  p1[1] <- (1 - f1) * p0
  f2 <- (1 / 2 + 3 / 2 * f1) / 2
  p1[2] <- (1 - f2) * p0
  ff2 <- (3 + 5 * f1) / 8 / n + (1 - 1 / n) / 4 * (1 + 3 * f1)
  f3 <- P[3] * (1 / 4 + 1 / 4 * f1) + f2 / 2 + (1 - P[3]) / 2 * ff2
  p1[3] <- (1 - f3) * p0
  ff3 <- P1[3] / 8 * (3 + 4 * f2 + f1) +
    P2[3] / 4 * (1 + 2 * f2 + ff2) +
    P3[3] / 4 * (2 * f2 + f1 + 1 / 2 + f1 / 2) +
    P4[3] / 4 * (2 * f2 + ff2 + f1)
  f4 <- P[4] * (1 / 4 + 1 / 4 * f2) + f3 / 2 + (1 - P[4]) / 2 * ff3
  p1[4] <- (1 - f4) * p0

  cal_coef_hd <- function(P, P_, P__, P1, P2, P3) {
    c(
      P / 4 + (1 - P) / 2 * (P1 * 3 / 8 + P2 / 4 + P3 / 8 -
        (1 - P1 - P3) / 2 / (1 - P_) / 4 * P_ -
        (1 - P1 - P2) / 2 / (1 - P__) / 4 * P__),
      1 / 2 + (1 - P) / 2 * (1 - P1 - P3) / 2 / (1 - P_),
      P / 4 + (1 - P) / 4 - (1 - P) / 2 * (1 - P1 - P3) / 4 / (1 - P_) +
        (1 - P) / 2 * (1 - P1 - P2) / 2 / (1 - P__),
      (1 - P) / 2 * (P1 / 8 + P3 / 8 - (1 - P1 - P3) * P_ / 8 / (1 - P_) -
        (1 - P1 - P2) / 4 / (1 - P__)),
      -(1 - P) / 2 * (1 - P1 - P2) * P__ / 8 / (1 - P__)
    )
  }
  for (t in 5:Tt) {
    c_hd <- cal_coef_hd(P[t], P[t - 1], P[t - 2], P1[t - 1], P2[t - 1], P3[t - 1])
    p1[t] <- c_hd[2] * p1[t - 1] + c_hd[3] * p1[t - 2] + c_hd[4] * p1[t - 3] + c_hd[5] * p1[t - 4]
  }

  # -----------------------------
  # Model 2: diploid arrhenotoky
  # -----------------------------
  p2 <- rep(0, Tt)
  p2[1] <- (1 - f1) * p0
  f2 <- (1 / 2 + 3 / 2 * f1) / 2
  p2[2] <- (1 - f2) * p0
  ff2 <- (1 + 3 * f1) / 4 / n + (1 + 7 * f1) / 8 * (1 - 1 / n)
  f3 <- P[3] * (1 / 4 + 1 / 4 * f1) + f2 / 2 + (1 - P[3]) / 2 * ff2
  p2[3] <- (1 - f3) * p0
  ff3 <- P1[3] / 4 * (1 + 2 * f2 + f1) +
    P2[3] / 4 * (1 / 2 + 2 * f2 + f1 / 2 + ff2) +
    P3[3] / 4 * (2 * f2 + f1 + 1 / 2 + f1 / 2) +
    P4[3] / 4 * (2 * f2 + ff2 + f1)
  f4 <- P[4] * (1 / 4 + 1 / 4 * f2) + f3 / 2 + (1 - P[4]) / 2 * ff3
  p2[4] <- (1 - f4) * p0

  cal_coef_da <- function(P, P_, P__, P1, P2, P3) {
    c(
      P / 4 + (1 - P) / 2 * (P1 / 4 + P2 / 8 + P3 / 8 -
        (1 - P1 - P3) / 2 / (1 - P_) / 4 * P_ -
        (1 - P1 - P2) / 2 / (1 - P__) / 4 * P__),
      1 / 2 + (1 - P) / 2 * (1 - P1 - P3) / 2 / (1 - P_),
      P / 4 + (1 - P) / 4 - (1 - P) / 2 * (1 - P1 - P3) / 4 / (1 - P_) +
        (1 - P) / 2 * (1 - P1 - P2) / 2 / (1 - P__),
      (1 - P) / 2 * (P1 / 8 + P3 / 8 - (1 - P1 - P3) * P_ / 8 / (1 - P_) -
        (1 - P1 - P2) / 4 / (1 - P__)),
      (1 - P) / 2 * (P1 / 8 + P2 / 8 - (1 - P1 - P2) * P__ / 8 / (1 - P__))
    )
  }
  for (t in 5:Tt) {
    c_da <- cal_coef_da(P[t], P[t - 1], P[t - 2], P1[t - 1], P2[t - 1], P3[t - 1])
    p2[t] <- c_da[2] * p2[t - 1] + c_da[3] * p2[t - 2] + c_da[4] * p2[t - 3] + c_da[5] * p2[t - 4]
  }

  # -----------------------------
  # Model 3: complete diploidy
  # -----------------------------
  p3 <- rep(0, Tt)
  p3[1] <- (1 - f1) * p0
  f2 <- (1 + 3 * f1) / 4 / n + (1 + 7 * f1) / 8 * (1 - 1 / n)
  p3[2] <- (1 - f2) * p0
  fm2 <- (1 / 2 + 7 / 2 * f1) / 4
  f3 <- P1[3] / 4 * (1 + 2 * f2 + f1) +
    P2[3] / 4 * (1 / 2 + 2 * f2 + f1 / 2 + f2) +
    P3[3] / 4 * (2 * f2 + 1 / 2 + f1 / 2) +
    P4[3] / 4 * (2 * f2 + f2 + fm2)
  p3[3] <- (1 - f3) * p0
  if (N[3] > 1) {
    fm3 <- f2 / 2 + f2 / 4 + P2[3] / (1 - P1[3] - P3[3]) / 4 * (1 / 2 + f1 / 2) +
      P4[3] / (1 - P1[3] - P3[3]) / 4 * fm2
    f4 <- P1[4] / 4 * (1 + 2 * f3 + f2) +
      P2[4] / 4 * (1 / 2 + 2 * f3 + f2 / 2 + f3) +
      P3[4] / 4 * (2 * f3 + 1 / 2 + f2 / 2) +
      P4[4] / 4 * (2 * f3 + f3 + fm3)
  } else {
    f4 <- P1[4] / 4 * (1 + 2 * f3 + f2) + P2[4] / 4 * (1 / 2 + 2 * f3 + f2 / 2 + f3)
  }
  p3[4] <- (1 - f4) * p0

  cal_coef_cd <- function(P1, P2, P3, P4, P1_, P2_, P3_, P4_) {
    c(
      P1 / 4 + P2 / 8 + P3 / 8 + (1 - P1 - P2) / 4 * (P2_ / (1 - P1_ - P3_) / 8 -
        P4_ / (1 - P1_ - P3_) / (1 - P1_ - P2_) * (P1_ / 4 + P2_ / 8 + P3_ / 8)),
      1 / 2 + (1 - P1 - P3) / 4 + (1 - P1 - P2) / 4 * P4_ / (1 - P1_ - P3_) / (1 - P1_ - P2_),
      P1 / 4 + P2 / 8 + P3 / 8 + (1 - P1 - P2) / 4 * (3 / 4 -
        P4_ / (1 - P1_ - P2_) / (1 - P1_ - P3_) * (1 / 2 + (1 - P1_ - P3_) / 4)),
      (1 - P1 - P2) / 4 * (P2_ / (1 - P1_ - P3_) / 8 -
        P4_ / (1 - P1_ - P3_) / (1 - P1_ - P2_) * (P1_ / 4 + P2_ / 8 + P3_ / 8))
    )
  }
  for (t in 5:Tt) {
    if (N[t - 2] > 1) {
      c_cd <- cal_coef_cd(P1[t], P2[t], P3[t], P4[t], P1[t - 1], P2[t - 1], P3[t - 1], P4[t - 1])
      p3[t] <- c_cd[2] * p3[t - 1] + c_cd[3] * p3[t - 2] + c_cd[4] * p3[t - 3]
    } else if (n > 1) {
      c_cd <- c(P1[t] / 4 + P2[t] / 8, P1[t] / 2 + 3 * P2[t] / 4, P1[t] / 4 + P2[t] / 8)
      p3[t] <- c_cd[2] * p3[t - 1] + c_cd[3] * p3[t - 2]
    } else {
      c_cd <- c(P1[t] / 4, P1[t] / 2, P1[t] / 4)
      p3[t] <- c_cd[2] * p3[t - 1] + c_cd[3] * p3[t - 2]
    }
  }

  # Derived trajectories
  gen <- 2:Tt
  loss_hd <- p1[-1] / p1[-Tt] - 1
  loss_da <- p2[-1] / p2[-Tt] - 1
  loss_cd <- p3[-1] / p3[-Tt] - 1
  ne_hd <- -1 / 2 / loss_hd
  ne_da <- -1 / 2 / loss_da
  ne_cd <- -1 / 2 / loss_cd

  list(
    model_df = tibble(
      generation = rep(gen, 3),
      model = factor(
        rep(c("Haplo-diploidy", "Diploid arrhenotoky", "Complete diploidy"), each = length(gen)),
        levels = c("Haplo-diploidy", "Diploid arrhenotoky", "Complete diploidy")
      ),
      heterozygosity_loss = c(loss_hd, loss_da, loss_cd),
      effective_ne = c(ne_hd, ne_da, ne_cd)
    ),
    saturation_gen = suppressWarnings(min(which(N == C))),
    C = C,
    K = K
  )
}

# Two scenarios with the same C/K ratio (C/K = 2)
scenario_low <- run_varroa_model(C = 20, K = 10)
scenario_high <- run_varroa_model(C = 200, K = 100)

# Keep these objects available for the heterozygosity chunk (panel B)
model_df <- scenario_high$model_df
model_colors <- c(
  "Haplo-diploidy" = "red3",
  "Diploid arrhenotoky" = "dodgerblue3",
  "Complete diploidy" = "black"
)

ne_panel_df <- bind_rows(
  scenario_low$model_df %>%
    filter(model != "Complete diploidy") %>%
    mutate(panel = "a"),
  scenario_high$model_df %>%
    filter(model != "Complete diploidy") %>%
    mutate(panel = "b")
) %>%
  mutate(panel = factor(panel, levels = c("a", "b")))

vline_df <- tibble(
  panel = factor(c("a", "b"), levels = c("a", "b")),
  saturation_gen = c(scenario_low$saturation_gen, scenario_high$saturation_gen)
)

panel_label_df <- tibble(
  panel = factor(c("a", "b"), levels = c("a", "b")),
  x = c(5.2, 5.2),
  y = c(14.6, 14.6),
  label = c("a", "b")
)

model_colors <- c(
  "Haplo-diploidy" = "red3",
  "Diploid arrhenotoky" = "dodgerblue3",
  "Complete diploidy" = "black"
)

# Plot 1A-B: Effective population size (Ne), two C/K scenarios
g_ne <- ggplot(
  ne_panel_df,
  aes(x = generation, y = effective_ne, color = model)
) +
  geom_line(linewidth = 1) +
  theme_classic() +
  scale_color_manual(values = model_colors) +
  coord_cartesian(xlim = c(5, 30), ylim = c(2, 15)) +
  facet_wrap(~panel, ncol = 2) +
  labs(
    x = "Generation",
    y = "Effective population size (Ne)",
    color = "Model"
  ) +
  geom_vline(
    data = vline_df %>% filter(is.finite(saturation_gen)),
    aes(xintercept = saturation_gen),
    inherit.aes = FALSE,
    linetype = "dashed",
    color = "gray40"
  ) +
  geom_text(
    data = panel_label_df,
    aes(x = x, y = y, label = label),
    inherit.aes = FALSE,
    fontface = "bold",
    size = 5
  ) +
  theme(
    legend.position = "bottom",
    panel.border = element_rect(color = "black", fill = NA, linewidth = 0.6),
    panel.spacing = unit(1.2, "lines"),
    strip.text = element_blank(),
    strip.background = element_blank()
  )

g_ne

cat("\n**Supplementary Figure 1.** Effective population size (Ne) across generations under haplo-diploidy (red) and diploid arrhenotoky (blue), shown for two model scenarios with the same C/K ratio (C/K = 2). **Panel (a):** C = 20, K = 10. **Panel (b):** C = 200, K = 100. Dashed vertical lines mark the generation at which the simulated population first reaches carrying capacity in each scenario.")

# save ne_panel_df for the Source data excel file
#write.csv(ne_panel_df,
#  file = file.path(source_dir, "SuppFig1_Ne_model_curves.csv"),
#  row.names = FALSE)
```

```{r model-heterozygosity, eval=FALSE, message=FALSE, warning=FALSE, include=FALSE}
# Relative heterozygosity loss rate per generation
g_loss <- ggplot(model_df, aes(x = generation, y = heterozygosity_loss, color = model)) +
  geom_line(linewidth = 1) +
  theme_classic() +
  scale_color_manual(values = model_colors) +
  labs(
    title = "B",
    x = "Generation",
    y = "Relative heterozygosity loss rate",
    color = "Model"
  ) +
  theme(legend.position = "bottom")

g_loss

cat("\n**Supplementary Figure 1B.** Relative heterozygosity loss rate per generation under haplo-diploidy (red), diploid arrhenotoky (blue), and complete diploidy (black).")
```


# Genomic Pedigree study
This analysis is part of a comprehensive study investigating the genetic
inheritance patterns in *Varroa destructor* mites. We analyzed genomic
data from 223 mite samples across 30 families, collected from three
*Apis mellifera* colonies. Each family consists of three generations: F0 foundress adult female mite, F1 male and female adult offspring mites, and their F2 offspring, composed of immature mites (nymphs and eggs) and at least one adult male.    

The analysis should be read in conjunction with the detailed methodology presented in [Site
Quality Analysis](https://github.com/nurit-eliash/varroa-pedigree-study/blob/main/R_scripts/After_revision/site_quality_exploration.Rmd).

### Data Availability

-   Raw sequencing data: [BioProject
    PRJNA794941](https://www.ncbi.nlm.nih.gov/bioproject/PRJNA794941/)
-   Analysis code: [GitHub
    Repository](https://github.com/nurit-eliash/varroa-pedigree-study/)
-   Filtered datasets: Available upon request

### Data Sources

-   Raw fastq reads: [BioProject
    PRJNA794941](https://www.ncbi.nlm.nih.gov/bioproject/PRJNA794941/)
-   Code repository:
    [GitHub](https://github.com/nurit-eliash/varroa-pedigree-study/tree/main#varroa-pedigree-study)

### Required Files

Before running this analysis, ensure you have:\
1. The VCF file processed with the specified parameters\
2. The filtered sites CSV file generated from site quality exploration\
3. All required R packages installed


## Data Processing

## VCF File Processing

The initial variant calling file was filtered using VCFtools with
stringent quality control parameters to ensure high-confidence variant
calls. The filtering criteria were:

``` bash
vcftools --vcf snponly_freebayes.vcf \
  --chr NW_019211454.1 \
  --chr NW_019211455.1 \
  --chr NW_019211456.1 \
  --chr NW_019211457.1 \
  --chr NW_019211458.1 \
  --chr NW_019211459.1 \
  --chr NW_019211460.1 \
  --max-alleles 2 \
  --minQ 15000 \
  --minDP 16 \
  --maxDP 40 \
  --max-missing 0.5 \
  --maf 0.2 \
  --recode \
  --recode-INFO-all \
  --out Q15000BIALLDP16HDP40mis.5maf.2Chr7
```

### VCF file filtering rationale

-   **Chromosomes**: Selected the 7 largest chromosomes
    (NW_019211454.1-460.1)
-   **Alleles**: Limited to biallelic sites to focus on clear binary
    inheritance patterns
-   **Quality**: Minimum Phred score of 15000 ensures high base quality
-   **Depth**: Range of 16-40 balances between coverage requirements and
    potential artifacts
-   **Missing Data**: Maximum 50% missing data per site maintains
    statistical power
-   **Minor Allele Frequency**: Minimum 0.2 reduces impact of sequencing
    errors

## Quality Control Metrics

We implemented three key metrics to assess genotype quality:

1.  **Variant Quality Distribution**       

2.  **Variant Allele Frequency (VAF)**
    -   Calculated as: VAF = AO/DP
    -   AO: Alternate allele observations
    -   DP: Total read depth
    -   Provides measure of allelic balance
3.  **Quality Allele Frequency (QAF)**
    -   Calculated as: QAF = QA/(QA+QO)
    -   QA: Quality scores for alternate allele
    -   QO: Quality scores for reference allele
    -   Weights allele calls by base quality

### Supplementary Figure 3: Variant Quality Distribution 
```{r variant-quality-dist, echo=TRUE, eval=TRUE, warning=FALSE, message=FALSE}
# Read the variant quality data
var_qual <- read_delim("/Users/nuriteliash/Documents/GitHub/varroa-pedigree-study/data/vcf_filter/Q40BIALLDP16HDP40mis.5Chr7/Q40BIALLDP16HDP40mis.5Chr7.lqual", 
                       delim = "\t",
                       col_names = c("chr", "pos", "qual"), 
                       skip = 1)

# Print summary statistics
cat("Summary of variant quality scores:\n")
print(summary(var_qual$qual))

# Create main plot
ggplot(var_qual, aes(x = qual)) +
  geom_density(fill = "dodgerblue1", colour = "black", alpha = 0.3) + 
    theme_classic() +
  ggtitle("Variant Quality Distribution") +
  scale_x_continuous(n.breaks = 12, limits = c(0, 50000)) +  # Adjusted x-axis range
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    plot.title = element_text(size = 12),
    axis.title = element_text(size = 10)
  ) +
  xlab("Variant Quality") + 
  ylab("Density") +
  geom_vline(xintercept = 15000, color = "red", linetype = "dashed")

# Calculate percentage of variants above threshold
pct_above_threshold <- mean(var_qual$qual >= 15000) * 100

# Add caption below the plot
cat("\n**Supplementary Figure 3.** Distribution of variant quality scores across sites that passed initial filtering (depth ≥16, ≤40; maximum 50% missing data). The red dashed line indicates the quality threshold (15000) used for final filtering. %.1f%% of variants have quality scores above this threshold. Higher quality scores indicate greater confidence in variant calls.", pct_above_threshold)

# save var_qual for the Source data excel file
#write.csv(var_qual,
#  file = file.path(source_dir, "SuppFig3_varqual_density.csv"),
#  row.names = FALSE)
```

### Genotype Classification Thresholds, VAF and QAF

Based on empirical distribution analysis, we
established the following thresholds:

| Genotype | VAF Range | QAF Range | Interpretation                 |
|----------|-----------|-----------|--------------------------------|
| AA (Ref) | \< 0.01   | \< 0.01   | Strong reference support       |
| AB (Het) | 0.2 - 0.8 | 0.2 - 0.8 | Balanced allele representation |
| BB (Alt) | \> 0.9    | \> 0.9    | Strong alternate support       |

These thresholds were validated using known inheritance patterns in our
pedigree structure.

```{r process-vcf, echo=TRUE, eval=TRUE}
# Read and process VCF file
vcf <- read.vcfR("/Users/nuriteliash/Documents/GitHub/varroa-pedigree-study/data/vcf_filter/minQ_filter/Q15000BIALLDP16HDP40mis.5maf.2Chr7.recode.vcf", verbose = FALSE)

# Extract genotype data
gt <- extract.gt(vcf, element = "GT") 
gt <- as.data.frame(t(gt)) %>%
  rownames_to_column("ID")

# Clean the ID names
gt$ID <- sub("_[^_]+$", "", gt$ID)

# Create the table object
table <- gt %>%
  t() %>%
  as.data.frame() %>%
  row_to_names(row_number = 1) %>%
  dplyr::select(contains(c("son", "dat", "fnd"))) %>%  # Keep only F0, F1, and F2 adults
  # Remove columns containing specific substrings
  dplyr::select(-matches("sis|479-2|565-2|535_1|535_3")) %>%
  # Convert GT values from numbers to letters
  mutate(across(everything(), ~ case_when(
    . == "0/0" ~ "AA",
    . == "0/1" ~ "AB",
    . == "1/1" ~ "BB",
    TRUE ~ .  # Keep other values unchanged
  )))
```

The site quality filtering criteria were developed based on Variant
Allele Frequency (VAF) and Genotype Quality Allele Frequency (QAF)
analysis. For detailed methodology and validation of these thresholds,
see [Site Quality
Analysis](site_quality_exploration.html#sanity-check-to-determine-the-threshold-for-site-filtering).

We applied two additional filtering parameters to account for biases in
alternate allele representation and to ensure high-confidence genotype
calls based on allele frequency and quality.\
*Variant allele frequency (VAF)* was calculated as VAF = AO/DP, where AO
(Alternate Observations) represents the number of reads supporting the
alternate allele, and DP (Depth) is the total read depth at the site.\
Similarly, *Quality Allele Frequency (QAF)* was computed as QAF =
QA/QA+QO, where QA (Quality of Alternate Allele) is the sum of
Phred-scaled base quality scores for reads supporting the alternate
allele, and QO (Quality of Reference Allele) is the sum of Phred-scaled
base quality scores for reads supporting the reference allele.

## Supplementary Figures 4-6 : Picking the thresholds for site quality: Distribution of VAF and QAF in F2 Offspring of homozygous parents

### Supplementary Figure 4, Quality Allele Frequency (QAF) and Variant Allele Frequency (VAF) in F2 females of AA parents. 
```{r load-f2-AA-parents}
# Read the f2_homo_parents data (sanity check data) from CSV: F2 females at the filtered positions, on chrom 1, where F1 male and female are homozygous (AA)
# the file was created using the ~/Documents/GitHub/varroa-pedigree-study/R_scripts/After_revision/site_quality_exploration.Rmd
f2_AA_parents <- read.csv("/Users/nuriteliash/Documents/GitHub/varroa-pedigree-study/data/f2_AA_parents.csv")
```

```{r plot-f2-AA-parents, echo=TRUE, eval=TRUE, fig.width=12}
# QAF log-scale plot
QAF_AA_gt_log <- ggplot(f2_AA_parents, aes(x = QAF, fill = GT)) +
  scale_x_continuous(breaks = seq(0, 1, by = 0.1), limits = c(-0.05, 1.05)) +
  scale_y_continuous(
    trans = "log10",
    breaks = 10^(0:5),
    labels = scales::trans_format("log10", scales::math_format(10^.x))
  ) +
  coord_cartesian(ylim = c(2, max(table(f2_AA_parents$QAF)) * 1.1)) +
  annotate("rect", xmin = -Inf, xmax = 0.01, ymin = 1, ymax = max(table(f2_AA_parents$QAF)) * 1.1, alpha = 0.2, fill = "#ffbf00") +
  geom_histogram(binwidth = 0.02, alpha = 1, position = "identity") +
  geom_vline(xintercept = c(0.01), linetype = "dashed", color = "gray", alpha = 1) +
  theme_classic() +
  labs(x = "Quality Allele Frequency (QAF)",
       y = "Count (log scale)",
       title = "A") +
  scale_fill_manual(
    values = c("AA" = "#ffbf00", "AB" = "#66b032", "BB" = "#1982c4"),
    name = "Genotype"
  ) +
  theme(
    legend.position = "none",
    axis.text.x = element_text(size = 10, angle = 45, hjust = 1, vjust = 1),
    axis.title.x = element_text(margin = margin(t = 15)),
    plot.margin = margin(10, 10, 30, 10),
    plot.title = element_text(face = "bold", hjust = -0.2)
  )

# VAF log-scale plot
VAF_AA_gt_log <- ggplot(f2_AA_parents, aes(x = VAF, fill = GT)) +
  scale_x_continuous(breaks = seq(0, 1, by = 0.1), limits = c(-0.05, 1.05)) +
  scale_y_continuous(
    trans = "log10",
    breaks = 10^(0:5),
    labels = scales::trans_format("log10", scales::math_format(10^.x))
  ) +
  coord_cartesian(ylim = c(2, max(table(f2_AA_parents$VAF)) * 1.1)) +
  annotate("rect", xmin = -Inf, xmax = 0.01, ymin = 1, ymax = max(table(f2_AA_parents$VAF)) * 1.1, alpha = 0.2, fill = "#ffbf00") +
  geom_histogram(binwidth = 0.02, alpha = 1, position = "identity") +
  geom_vline(xintercept = c(0.01), linetype = "dashed", color = "gray", alpha = 1) +
  theme_classic() +
  labs(x = "Variant Allele Frequency (VAF)",
       y = "Count (log scale)",
       title = "B") +
  scale_fill_manual(
    values = c("AA" = "#ffbf00", "AB" = "#66b032", "BB" = "#1982c4"),
    name = "Genotype"
  ) +
  theme(
    legend.position = "right",
    axis.text.x = element_text(size = 10, angle = 45, hjust = 1, vjust = 1),
    axis.title.x = element_text(margin = margin(t = 15)),
    plot.margin = margin(10, 10, 30, 10),
    plot.title = element_text(face = "bold", hjust = -0.2)
  )

# QAF non-log scale plot
QAF_AA_gt <- ggplot(f2_AA_parents, aes(x = QAF, fill = GT)) +
  scale_x_continuous(breaks = seq(0, 1, by = 0.1), limits = c(-0.05, 1.05)) +
  annotate("rect", xmin = -Inf, xmax = 0.01, ymin = 0, ymax = max(table(f2_AA_parents$QAF)) * 1.1, alpha = 0.2, fill = "#ffbf00") +
  geom_histogram(binwidth = 0.02, alpha = 1, position = "identity") +
  geom_vline(xintercept = c(0.01), linetype = "dashed", color = "gray", alpha = 1) +
  theme_classic() +
  labs(x = "Quality Allele Frequency (QAF)",
       y = "Count",
       title = "C") +
  scale_fill_manual(
    values = c("AA" = "#ffbf00", "AB" = "#66b032", "BB" = "#1982c4"),
    name = "Genotype"
  ) +
  theme(
    legend.position = "none",
    axis.text.x = element_text(size = 10, angle = 45, hjust = 1, vjust = 1),
    axis.title.x = element_text(margin = margin(t = 15)),
    plot.margin = margin(10, 10, 30, 10),
    plot.title = element_text(face = "bold", hjust = -0.2)
  )

# VAF non-log scale plot
VAF_AA_gt <- ggplot(f2_AA_parents, aes(x = VAF, fill = GT)) +
  scale_x_continuous(breaks = seq(0, 1, by = 0.1), limits = c(-0.05, 1.05)) +
  annotate("rect", xmin = -Inf, xmax = 0.01, ymin = 0, ymax = max(table(f2_AA_parents$VAF)) * 1.1, alpha = 0.2, fill = "#ffbf00") +
  geom_histogram(binwidth = 0.02, alpha = 1, position = "identity") +
  geom_vline(xintercept = c(0.01), linetype = "dashed", color = "gray", alpha = 1) +
  theme_classic() +
  labs(x = "Variant Allele Frequency (VAF)",
       y = "Count",
       title = "D") +
  scale_fill_manual(
    values = c("AA" = "#ffbf00", "AB" = "#66b032", "BB" = "#1982c4"),
    name = "Genotype"
  ) +
  theme(
    legend.position = "right",
    axis.text.x = element_text(size = 10, angle = 45, hjust = 1, vjust = 1),
    axis.title.x = element_text(margin = margin(t = 15)),
    plot.margin = margin(10, 10, 30, 10),
    plot.title = element_text(face = "bold", hjust = -0.2)
  )

# Combine plots with shared legend
combined_plot <- (QAF_AA_gt_log + VAF_AA_gt_log + QAF_AA_gt + VAF_AA_gt) + plot_layout(ncol = 4, guides = "collect")

# Display the combined plot
combined_plot

# Add unified caption below the plot
cat("\n**Figure S4.** Distribution of Quality Allele Frequency (QAF) and Variant Allele Frequency (VAF) values in F2 females from crosses with homozygous AA parents. **A-B.** QAF and VAF distributions shown on logarithmic scales. **C-D.** Same distributions shown on linear scales. In all plots, shaded regions indicate the thresholds for genotype classification: AA (yellow, < 0.01). Vertical dashed lines mark the threshold boundaries. The plots show the distribution of genotype calls, with AA genotypes predominantly falling below the 0.01 threshold range.")

# save f2_AA_parents for the Source data excel file
#write.csv(f2_AA_parents,
#          file = file.path(source_dir, "SuppFig4_QAF_VAF_F2_AA.csv"),
#          row.names = FALSE)
```

### Supplementary Figure 5, Quality Allele Frequency (QAF) and Variant Allele Frequency (VAF) in F2 females of BB parents. 
```{r load-f2-BB-parents}
# Read the f2_homo_parents data (sanity check data) from CSV: F2 females at the filtered positions, on chrom 1, where F1 male and female are homozygous (AA)
# the file was created using the ~/Documents/GitHub/varroa-pedigree-study/R_scripts/After_revision/site_quality_exploration.Rmd
f2_BB_parents <- read.csv("/Users/nuriteliash/Documents/GitHub/varroa-pedigree-study/data/f2_BB_parents.csv")
```

```{r plot-f2-BB-parents, echo=TRUE, eval=TRUE, fig.width=12}
# QAF log-scale plot
QAF_BB_gt_log <- ggplot(f2_BB_parents, aes(x = QAF, fill = GT)) +
  scale_x_continuous(breaks = seq(0, 1, by = 0.1), limits = c(-0.05, 1.05)) +
  scale_y_continuous(
    trans = "log10",
    breaks = 10^(0:5),
    labels = scales::trans_format("log10", scales::math_format(10^.x))
  ) +
  coord_cartesian(ylim = c(2, max(table(f2_BB_parents$QAF)) * 1.1)) +
  annotate("rect", xmin = 0.9, xmax = Inf, ymin = 1, ymax = max(table(f2_BB_parents$QAF)) * 1.1, alpha = 0.2, fill = "#1982c4") +
  geom_histogram(binwidth = 0.02, alpha = 1, position = "identity") +
  geom_vline(xintercept = c(0.9), linetype = "dashed", color = "gray", alpha = 1) +
  theme_classic() +
  labs(x = "Quality Allele Frequency (QAF)",
       y = "Count (log scale)",
       title = "A") +
  scale_fill_manual(
    values = c("AA" = "#ffbf00", "AB" = "#66b032", "BB" = "#1982c4"),
    name = "Genotype"
  ) +
  theme(
    legend.position = "none",
    axis.text.x = element_text(size = 10, angle = 45, hjust = 1, vjust = 1),
    axis.title.x = element_text(margin = margin(t = 15)),
    plot.margin = margin(10, 10, 30, 10),
    plot.title = element_text(face = "bold", hjust = -0.2)
  )

# VAF log-scale plot
VAF_BB_gt_log <- ggplot(f2_BB_parents, aes(x = VAF, fill = GT)) +
  scale_x_continuous(breaks = seq(0, 1, by = 0.1), limits = c(-0.05, 1.05)) +
  scale_y_continuous(
    trans = "log10",
    breaks = 10^(0:5),
    labels = scales::trans_format("log10", scales::math_format(10^.x))
  ) +
  coord_cartesian(ylim = c(2, max(table(f2_BB_parents$VAF)) * 1.1)) +
  annotate("rect", xmin = 0.9, xmax = Inf, ymin = 1, ymax = max(table(f2_BB_parents$VAF)) * 1.1, alpha = 0.2, fill = "#1982c4") +
  geom_histogram(binwidth = 0.02, alpha = 1, position = "identity") +
  geom_vline(xintercept = c(0.9), linetype = "dashed", color = "gray", alpha = 1) +
  theme_classic() +
  labs(x = "Variant Allele Frequency (VAF)",
       y = "Count (log scale)",
       title = "B") +
  scale_fill_manual(
    values = c("AA" = "#ffbf00", "AB" = "#66b032", "BB" = "#1982c4"),
    name = "Genotype"
  ) +
  theme(
    legend.position = "right",
    axis.text.x = element_text(size = 10, angle = 45, hjust = 1, vjust = 1),
    axis.title.x = element_text(margin = margin(t = 15)),
    plot.margin = margin(10, 10, 30, 10),
    plot.title = element_text(face = "bold", hjust = -0.2)
  )

# QAF non-log scale plot
QAF_BB_gt <- ggplot(f2_BB_parents, aes(x = QAF, fill = GT)) +
  scale_x_continuous(breaks = seq(0, 1, by = 0.1), limits = c(-0.05, 1.05)) +
  annotate("rect", xmin = 0.9, xmax = Inf, ymin = 0, ymax = max(table(f2_BB_parents$QAF)) * 1.1, alpha = 0.2, fill = "#1982c4") +
  geom_histogram(binwidth = 0.02, alpha = 1, position = "identity") +
  geom_vline(xintercept = c(0.9), linetype = "dashed", color = "gray", alpha = 1) +
  theme_classic() +
  labs(x = "Quality Allele Frequency (QAF)",
       y = "Count",
       title = "C") +
  scale_fill_manual(
    values = c("AA" = "#ffbf00", "AB" = "#66b032", "BB" = "#1982c4"),
    name = "Genotype"
  ) +
  theme(
    legend.position = "none",
    axis.text.x = element_text(size = 10, angle = 45, hjust = 1, vjust = 1),
    axis.title.x = element_text(margin = margin(t = 15)),
    plot.margin = margin(10, 10, 30, 10),
    plot.title = element_text(face = "bold", hjust = -0.2)
  )

# VAF non-log scale plot
VAF_BB_gt <- ggplot(f2_BB_parents, aes(x = VAF, fill = GT)) +
  scale_x_continuous(breaks = seq(0, 1, by = 0.1), limits = c(-0.05, 1.05)) +
  annotate("rect", xmin = 0.9, xmax = Inf, ymin = 0, ymax = max(table(f2_BB_parents$VAF)) * 1.1, alpha = 0.2, fill = "#1982c4") +
  geom_histogram(binwidth = 0.02, alpha = 1, position = "identity") +
  geom_vline(xintercept = c(0.9), linetype = "dashed", color = "gray", alpha = 1) +
  theme_classic() +
  labs(x = "Variant Allele Frequency (VAF)",
       y = "Count",
       title = "D") +
  scale_fill_manual(
    values = c("AA" = "#ffbf00", "AB" = "#66b032", "BB" = "#1982c4"),
    name = "Genotype"
  ) +
  theme(
    legend.position = "right",
    axis.text.x = element_text(size = 10, angle = 45, hjust = 1, vjust = 1),
    axis.title.x = element_text(margin = margin(t = 15)),
    plot.margin = margin(10, 10, 30, 10),
    plot.title = element_text(face = "bold", hjust = -0.2)
  )

# Combine plots with shared legend
combined_plot <- (QAF_BB_gt_log + VAF_BB_gt_log + QAF_BB_gt + VAF_BB_gt) + plot_layout(ncol = 4, guides = "collect")

# Display the combined plot
combined_plot

# Add unified caption below the plot
cat("\n**Figure S5.** Distribution of Quality Allele Frequency (QAF) and Variant Allele Frequency (VAF) values in F2 females from crosses with homozygous BB parents. **A-B.** QAF and VAF distributions shown on logarithmic scales. **C-D.** Same distributions shown on linear scales. In all plots, shaded regions indicate the thresholds for genotype classification: BB (blue, > 0.9). Vertical dashed lines mark the threshold boundaries. The plots show the distribution of genotype calls, with BB genotypes predominantly falling above the 0.9 threshold range.")

# save f2_BB_parents for the Source data excel file
#write.csv(f2_BB_parents,
#          file = file.path(source_dir, "SuppFig5_QAF_VAF_F2_BB.csv"),
#          row.names = FALSE)
```

### Supplementary Figure 6, Quality Allele Frequency (QAF) and Variant Allele Frequency (VAF) in F2 females of AA and AB parents. 

```{r load-f2-AB-parents}
# Read the f2_homo_parents data (sanity check data) from CSV: F2 females at the filtered positions, on chrom 1, where F1 male and female are homozygous (AA)
# the file was created using the ~/Documents/GitHub/varroa-pedigree-study/R_scripts/After_revision/site_quality_exploration.Rmd
f2_AB_combined <- read.csv("/Users/nuriteliash/Documents/GitHub/varroa-pedigree-study/data/f2_AB_combined.csv")
```

```{r plot-f2-AB-parents, eval=TRUE, fig.width=12}
# QAF log-scale plot
QAF_AB_gt_log <- ggplot(f2_AB_combined, aes(x = QAF, fill = GT)) +
  # First set up the basic plot structure and coordinates
  scale_x_continuous(breaks = seq(0, 1, by = 0.1), limits = c(-0.05, 1.05)) +
  scale_y_continuous(
    trans = "log10",
    breaks = 10^(0:5),
    labels = scales::trans_format("log10", scales::math_format(10^.x))
  ) +
  coord_cartesian(ylim = c(2, max(table(f2_AB_combined$QAF)) * 1.1)) +
  # Add rectangles with explicit y-limits matching the coordinate system
  annotate("rect", xmin = 0.2, xmax = 0.8, ymin = 1, ymax = max(table(f2_AB_combined$QAF)) * 1.1, alpha = 0.2, fill = "#66b032") +
  # Add the data layers
  geom_histogram(binwidth = 0.02, alpha = 1, position = "identity") +
  geom_vline(xintercept = c(0.2, 0.8), linetype = "dashed", color = "gray", alpha = 1) +
  # Theme and labels
  theme_classic() +
  labs(x = "Quality Allele Frequency (QAF)",
       y = "Count (log scale)",
       title = "A") +
  scale_fill_manual(
    values = c("AA" = "#ffbf00", "AB" = "#66b032", "BB" = "#1982c4"),
    name = "Genotype"
  ) +
  theme(
    legend.position = "none",
    axis.text.x = element_text(size = 10, angle = 45, hjust = 1, vjust = 1),
    axis.title.x = element_text(margin = margin(t = 15)),
    plot.margin = margin(10, 10, 30, 10),
    plot.title = element_text(face = "bold", hjust = -0.2)
  )

# VAF log-scale plot
VAF_AB_gt_log <- ggplot(f2_AB_combined, aes(x = VAF, fill = GT)) +
  # First set up the basic plot structure and coordinates
  scale_x_continuous(breaks = seq(0, 1, by = 0.1), limits = c(-0.05, 1.05)) +
  scale_y_continuous(
    trans = "log10",
    breaks = 10^(0:5),
    labels = scales::trans_format("log10", scales::math_format(10^.x))
  ) +
  coord_cartesian(ylim = c(2, max(table(f2_AB_combined$VAF)) * 1.1)) +
  # Add rectangles with explicit y-limits matching the coordinate system
  annotate("rect", xmin = 0.2, xmax = 0.8, ymin = 1, ymax = max(table(f2_AB_combined$VAF)) * 1.1, alpha = 0.2, fill = "#66b032") +
  # Add the data layers
  geom_histogram(binwidth = 0.02, alpha = 1, position = "identity") +
  geom_vline(xintercept = c(0.2, 0.8), linetype = "dashed", color = "gray", alpha = 1) +
  # Theme and labels
  theme_classic() +
  labs(x = "Variant Allele Frequency (VAF)",
       y = "Count (log scale)",
       title = "B") +
  scale_fill_manual(
    values = c("AA" = "#ffbf00", "AB" = "#66b032", "BB" = "#1982c4"),
    name = "Genotype"
  ) +
  theme(
    legend.position = "right",
    axis.text.x = element_text(size = 10, angle = 45, hjust = 1, vjust = 1),
    axis.title.x = element_text(margin = margin(t = 15)),
    plot.margin = margin(10, 10, 30, 10),
    plot.title = element_text(face = "bold", hjust = -0.2)
  )

# QAF non-log scale plot
QAF_AB_gt <- ggplot(f2_AB_combined, aes(x = QAF, fill = GT)) +
  # First set up the basic plot structure and coordinates
  scale_x_continuous(breaks = seq(0, 1, by = 0.1), limits = c(-0.05, 1.05)) +
  # Add rectangles with explicit y-limits matching the coordinate system
  annotate("rect", xmin = 0.2, xmax = 0.8, ymin = 0, ymax = max(table(f2_AB_combined$QAF)) * 1.1, alpha = 0.2, fill = "#66b032") +
  # Add the data layers
  geom_histogram(binwidth = 0.02, alpha = 1, position = "identity") +
  geom_vline(xintercept = c(0.2, 0.8), linetype = "dashed", color = "gray", alpha = 1) +
  # Theme and labels
  theme_classic() +
  labs(x = "Quality Allele Frequency (QAF)",
       y = "Count",
       title = "C") +
  scale_fill_manual(
    values = c("AA" = "#ffbf00", "AB" = "#66b032", "BB" = "#1982c4"),
    name = "Genotype"
  ) +
  theme(
    legend.position = "none",
    axis.text.x = element_text(size = 10, angle = 45, hjust = 1, vjust = 1),
    axis.title.x = element_text(margin = margin(t = 15)),
    plot.margin = margin(10, 10, 30, 10),
    plot.title = element_text(face = "bold", hjust = -0.2)
  )

# VAF non-log scale plot
VAF_AB_gt <- ggplot(f2_AB_combined, aes(x = VAF, fill = GT)) +
  # First set up the basic plot structure and coordinates
  scale_x_continuous(breaks = seq(0, 1, by = 0.1), limits = c(-0.05, 1.05)) +
  # Add rectangles with explicit y-limits matching the coordinate system
  annotate("rect", xmin = 0.2, xmax = 0.8, ymin = 0, ymax = max(table(f2_AB_combined$VAF)) * 1.1, alpha = 0.2, fill = "#66b032") +
  # Add the data layers
  geom_histogram(binwidth = 0.02, alpha = 1, position = "identity") +
  geom_vline(xintercept = c(0.2, 0.8), linetype = "dashed", color = "gray", alpha = 1) +
  # Theme and labels
  theme_classic() +
  labs(x = "Variant Allele Frequency (VAF)",
       y = "Count",
       title = "D") +
  scale_fill_manual(
    values = c("AA" = "#ffbf00", "AB" = "#66b032", "BB" = "#1982c4"),
    name = "Genotype"
  ) +
  theme(
    legend.position = "right",
    axis.text.x = element_text(size = 10, angle = 45, hjust = 1, vjust = 1),
    axis.title.x = element_text(margin = margin(t = 15)),
    plot.margin = margin(10, 10, 30, 10),
    plot.title = element_text(face = "bold", hjust = -0.2)
  )

# Combine plots with shared legend
combined_plot <- (QAF_AB_gt_log + VAF_AB_gt_log) / (QAF_AB_gt + VAF_AB_gt) + plot_layout(ncol = 2, guides = "collect")

# Display the combined plot
combined_plot

# Add unified caption below the plot
cat("\n**Supplementary Figure 6.** Distribution of Quality Allele Frequency (QAF) and Variant Allele Frequency (VAF) values in F2 females from crosses with heterozygous AB parents. **A-B.** QAF and VAF distributions shown on logarithmic scales. **C-D.** Same distributions shown on linear scales. In all plots, shaded regions indicate the thresholds for genotype classification: AB (green, 0.2-0.8). Vertical dashed lines mark the threshold boundaries. The plots show the distribution of genotype calls, with AB genotypes predominantly falling within the 0.2-0.8 threshold range.")

# save f2_AB_combined for the Source data excel file
#write.csv(f2_AB_combined,
#          file = file.path(source_dir, "SuppFig6_QAF_VAF_F2_AB.csv"),
#          row.names = FALSE)

```



Based on the distribution of sites' VAF and QAF values, the following
filtering criteria were applied:\
- Homozygous Reference (AA): VAF \< 0.01 and QAF \< 0.01, sites exhibit
low alternate allele support\
- Heterozygous (AB): 0.2 \< VAF \< 0.8 and 0.2 \< QAF \< 0.8, sites
maintain a balanced allele distribution\
- Homozygous Alternate (BB): VAF \> 0.9 and QAF \> 0.9, sites
demonstrate high alternate allele support

This filtering approach enhances the accuracy of genotype classification
while reducing the inclusion of low-confidence variant calls, by
minimizing genotyping errors due to sequencing artifacts or
depth-related biases.

## set the thresholds for QAF and VAF
```{r set-thresholds-QAF-VAF}
all_param <- readRDS("/Users/nuriteliash/Documents/GitHub/varroa-pedigree-study/data/all_param.rds")

# Create the filtered sites dataset that will be used in further analsyis
filtered_sites <- all_param %>%
  filter(
    (GT == "AA" & QAF < 0.01 & VAF < 0.01) |   # Homozygous Reference (AA)
    (GT == "AB" & QAF > 0.2 & QAF < 0.8 & VAF > 0.2 & VAF < 0.8) |  # Heterozygous (AB)
    (GT == "BB" & QAF > 0.9 & VAF > 0.9)    # Homozygous Alternate (BB)
  ) %>%
  select(family, Sample, site, QAF, VAF, GT)   %>%
   mutate(family = as.character(family),
Sample = as.character(Sample), # Extract Chromosome from the site column
chrom = str_split(site, pattern = "_", simplify = TRUE)[, 2]) %>% 
rename(gt = GT) # Keep only necessary columns
```

## Main figures

## Figure 1A: Crossing F1 mother AB with F1 father AA, one family

`~/Documents/GitHub/varroa-pedigree-study/R_scripts/After_revision/site_quality_exploration.Rmd`

```{r load-selected-family-478}
# Load the fam2_selected data, for one family (478) on the largest chromosome, NW_019211454
fam_478_F0AA_F1AB <- read.csv("/Users/nuriteliash/Documents/GitHub/varroa-pedigree-study/data/fam_478_F0AA_F1AB.csv")
```

```{r plot-one-family-478}
# Create a numeric x-axis for spacing (convert 'member' to factor and assign spacing)
fam_id <- 478

fam_478_F0AA_F1AB %>%
  ggplot(aes(x = member_num, y = pos / 1e6, color = GT)) +  # Convert pos to Mb
  geom_segment(aes(x = member_num - 0.3, xend = member_num + 0.3, 
                   y = pos / 1e6, yend = pos / 1e6), size = 1.5) +  # Apply same conversion
  scale_x_continuous(breaks = unique(fam_478_F0AA_F1AB$member_num), labels = unique(fam_478_F0AA_F1AB$member)) +  # Restore original labels
  scale_color_manual(
    values = c("AA" = "#ffbf00", "AB" = "#66b032", "BB" = "#1982c4"),  # Updated colors for renamed genotypes
    name = "Genotype",
    breaks = c("AA", "AB", "BB")  # Ensures BB appears in the legend even if absent
  ) +
  theme_classic() +
  labs(
    title = paste("Genotype Flow Across Family Members\nFamily ID:", fam_id),
    x = "Family Member",
    y = "Genomic Position (Mb)"
  ) +
  theme(
    axis.text.x = element_text(size = 10, angle = 45, hjust = 1, vjust = 1),  # Adjusted text size and angle
    axis.title.x = element_text(margin = margin(t = 15)),  # Add margin above x-axis title
    panel.grid.major = element_blank(), 
    legend.position = "right",  # Move legend to the right
    legend.box.margin = margin(0, 0, 0, 10),  # Add margin to separate legend from plot
    plot.margin = margin(10, 20, 30, 10)  # Increased bottom margin to accommodate rotated labels
  )

# Add caption below the plot
cat("\n**Figure 1A.** Genotype flow visualization across a single family (Family ID: 478). Each horizontal line represents a genomic position on chromosome NW_019211454.1, with colors indicating genotypes (yellow = AA, green = AB, blue = BB). Family members are arranged in generational order, showing the inheritance pattern from parents to offspring. The y-axis shows the genomic position in megabases (Mb).")

# save fam_478_F0AA_F1AB for the Source data excel file
#write.csv(fam_478_F0AA_F1AB,
#          file = file.path(source_dir, "Fig1A_single_family_tracks.csv"),
#          row.names = FALSE)
```

## Figure 1C: Crossing F1 mother AB with F1 father AA, all families

```{r load-and-process-ABxAA-pooled}
# Set the families (include only families with at least one adult F2)
family <- grep("grndat|grnson", gt$ID, value=TRUE) %>%
  str_extract("[^_]+") %>%
  unique()

# Define a list to store all data frames per family
obs <- list()

for (fam in family) {
    obs[[fam]] <- table %>%
      dplyr::select(starts_with(fam)) %>%
    # Add site and chromosome information
      rownames_to_column("site") %>%
    mutate(chrom = str_split(site, pattern = "_", simplify = TRUE)[, 2]) %>%
    # Then filter for the genotype patterns we want
      dplyr::filter_at(vars(matches("_fnd")), all_vars(. == "AA")) %>%  # F0 Female must be AA
      dplyr::filter_at(vars(matches("_son")), all_vars(. == "AA")) %>%  # Sons must be AA
      dplyr::filter_at(vars(matches("_dat")), all_vars(. == "AB")) %>%  # Daughters must be AB
    # Select only the grandchildren columns and keep site/chrom info
    dplyr::select(site, chrom, contains("grn")) %>%
    # Reshape the data
      tidyr::pivot_longer(cols = -c(site, chrom)) %>%
    dplyr::rename(Sample = name, gt = value) %>%
    # filter for sites 
    left_join(filtered_sites, by = c("site", "chrom", "Sample", "gt")) %>%
    na.omit() %>% # Count genotypes
    dplyr::count(Sample, gt, chrom, .drop = FALSE) %>%
      dplyr::filter(gt %in% c("AA", "BB", "AB")) %>%
      mutate(n = as.numeric(n)) %>%
    group_by(Sample) %>%
      mutate(total = as.numeric(sum(n))) %>%
      mutate(sex = case_when(
      grepl("son", Sample) ~ "Male",
      grepl("dat", Sample) ~ "Female"
      )) %>%
    # Add family column
      mutate(family = fam)
}

# Bind all families together into a final data frame containing all observed counts
observed <- do.call("rbind", obs) %>% 
  mutate(Sample = as.character(Sample))

# Create samples dataframe
samples <- data.frame(
  Sample = rep(unique(observed$Sample), each = 3), 
  gt = rep(c("AA", "AB", "BB"))
) %>%   
  mutate(family = str_extract(Sample, "^[0-9]+"))

# Join and create final dataset
samples_obs_filtered_ABxAA <- left_join(samples, observed, by = c("family", "Sample", "gt")) %>% 
  mutate(sex = case_when(
    grepl("son", Sample) ~ "Male",
    grepl("dat", Sample) ~ "Female"
  )) %>%
  group_by(Sample, chrom) %>%
    mutate(total = as.numeric(sum(n))) %>%
    dplyr::mutate(prop = n/total) %>%
    na.omit()

### Summary of sites filtered in each sample
# Create a table showing sites filtered per sample
sites_per_sample <- table %>%
  rownames_to_column("site") %>%
  tidyr::pivot_longer(cols = -site, names_to = "Sample", values_to = "gt") %>%
  # Filter out NA values before counting
  filter(!is.na(gt)) %>%
  group_by(Sample) %>%
  summarise(
    total_sites = n(),
    .groups = 'drop'
  )

filtered_sites_per_sample <- filtered_sites %>%
  group_by(Sample) %>%
  summarise(
    filtered_sites = n(),
    .groups = 'drop'
  )

# Join and calculate differences
sites_comparison <- sites_per_sample %>%
  left_join(filtered_sites_per_sample, by = "Sample") %>%
  mutate(
    filtered_sites = replace_na(filtered_sites, 0),
    sites_removed = total_sites - filtered_sites,
    percent_removed = round((sites_removed / total_sites) * 100, 2)
  ) %>%
  arrange(desc(sites_removed))

# Display the table
sites_comparison %>%
  select(Sample, total_sites, filtered_sites, sites_removed, percent_removed) %>%
  rename(
    "Sample" = Sample,
    "Total Sites" = total_sites,
    "Sites Kept" = filtered_sites,
    "Sites Removed" = sites_removed,
    "% Removed" = percent_removed
  ) %>%
  knitr::kable(
    digits = 2,
    caption = "Number of sites filtered per sample",
    format = "html"
  ) %>%
  kableExtra::kable_styling(
    bootstrap_options = "striped",
    full_width = TRUE,
    position = "left",
    fixed_thead = TRUE
  ) %>%
  kableExtra::scroll_box(
    width = "100%", 
    height = "300px"
  )

```

```{r plot-ABxAA-pooled, echo=TRUE, eval=TRUE, message=FALSE}
# Set minimum sites threshold
site_threshold = as.numeric(10)

# Create summary statistics for filtered data
summary_df_filtered <- samples_obs_filtered_ABxAA %>%
  filter(total >= site_threshold) %>%
  group_by(sex) %>%
  summarise(
    num_samples = n_distinct(Sample),  # Changed to Sample with capital S
    total_sites = sum(total),
    .groups = "drop"
  )

# Create summary statistics for filtered data by chromosome
summary_df_filtered_by_chrom <- samples_obs_filtered_ABxAA %>%
  filter(total >= site_threshold) %>%
  group_by(sex, chrom) %>%
  summarise(
    num_samples = n_distinct(Sample),  # Changed to Sample with capital S
    total_sites = sum(total),
    .groups = "drop"
  )

# Pulled data with totals across all chromosomes
ggplot(samples_obs_filtered_ABxAA %>% filter(total >= site_threshold), 
        aes(fill = gt, y = prop, x = factor(sex, levels = c("Female", "Male")))) +
  geom_bar(position = "fill", stat = "identity", width = 0.8) +
  ylab("Genotype Proportion") +
  labs(title = "Offspring Genotype, Crossing (AB) Male x (AA) Female
  Filtered sites", 
        fill = "Genotype") +
  theme_classic() +
  theme(
    axis.title.x = element_blank(),
    axis.text.x = element_blank(),
    axis.ticks.x = element_blank(),
    axis.text.y = element_text(size = 14),
    strip.text.x = element_text(size = 12),
    panel.spacing = unit(1, "lines"),
    plot.margin = margin(t = 10, r = 10, b = 0, l = 10)
  ) +
  scale_fill_manual(values = c("#ffbf00", "#66b032", "#1982c4")) +
  # Add sample count labels
  geom_text(data = summary_df_filtered,
            aes(x = sex, 
                y = -0.03,
                label = paste0(num_samples, " ", sex, "s")),
            inherit.aes = FALSE,
            size = 2.8,
            vjust = 1) +
  # Add sites count labels
  geom_text(data = summary_df_filtered,
            aes(x = sex, 
                y = -0.18,
                label = paste0(format(total_sites, big.mark = ","), " sites")),
            inherit.aes = FALSE,
            size = 2.8,
            vjust = 0.5) 

 # save var_qual for the Source data excel file
#write.csv(samples_obs_filtered_ABxAA,
#          file = file.path(source_dir, "Fig1C_ABxAA_barplot.csv"),
#          row.names = FALSE)

# save summary_df_filtered for the Source data excel file
#write.csv(summary_df_filtered,
#          file = file.path(source_dir, "Fig1C_sites_per_sample.csv"),
#          row.names = FALSE)
```

```{r fig1C-clean plot-no lables, echo=TRUE, eval=TRUE, message=FALSE}
# clean plot without labels , for the MS
ggplot(samples_obs_filtered_ABxAA %>% filter(total >= 10), 
       aes(fill = gt, y = prop, x = factor(sex, levels = c("Female", "Male")))) +
  geom_bar(position = "fill", stat = "identity", width = 0.7) +
  ylab("Genotype
Proportion") +
  labs(title = "Offspring Genotype, Crossing (AA) Male x (AB) Female
 Filtered sites", 
       fill = "Genotype") +
  theme_classic() +
  theme(
    axis.title.x = element_blank(),
    axis.text.x = element_blank(),
    axis.ticks.x = element_blank(),
    axis.text.y = element_text(size = 14),
    axis.title.y = element_text(size = 16),
    strip.text.x = element_text(size = 12),
    panel.spacing = unit(1, "lines"),
    plot.margin = margin(t = 10, r = 10, b = 0, l = 10)
  ) +
  scale_fill_manual(values = c("#ffbf00", "#66b032", "#1982c4"))# +
  #facet_wrap(~family)
```

## Figure 1B: Crossing F1 mother AA with F1 father AB, one family
```{r load-selected-family-534}
# Load the fam2_selected data, for one family (534) on the largest chromosome, NW_019211454
fam_534_F0AA_F1AA <- read.csv("/Users/nuriteliash/Documents/GitHub/varroa-pedigree-study/data/fam_534_F0AA_F1AA.csv")
```

```{r plot-one-family-534}
# Create a numeric x-axis for spacing (convert 'member' to factor and assign spacing)
fam_id <- 534

fam_534_F0AA_F1AA %>%
  ggplot(aes(x = member_num, y = pos / 1e6, color = GT)) +  # Convert pos to Mb
  geom_segment(aes(x = member_num - 0.3, xend = member_num + 0.3, 
                   y = pos / 1e6, yend = pos / 1e6), size = 1.5) +  # Apply same conversion
  scale_x_continuous(breaks = unique(fam_534_F0AA_F1AA$member_num), labels = unique(fam_534_F0AA_F1AA$member)) +  # Restore original labels
  scale_color_manual(
    values = c("AA" = "#ffbf00", "AB" = "#66b032", "BB" = "#1982c4"),  # Updated colors for renamed genotypes
    name = "Genotype",
    breaks = c("AA", "AB", "BB")  # Ensures BB appears in the legend even if absent
  ) +
  theme_classic() +
  labs(
    title = paste("Genotype Flow Across Family Members\nFamily ID:", fam_id),
    x = "Family Member",
    y = "Genomic Position (Mb)"
  ) +
  theme(
    axis.text.x = element_text(size = 10, angle = 45, hjust = 1, vjust = 1),  # Adjusted text size and angle
    axis.title.x = element_text(margin = margin(t = 15)),  # Add margin above x-axis title
    panel.grid.major = element_blank(), 
    legend.position = "right",  # Move legend to the right
    legend.box.margin = margin(0, 0, 0, 10),  # Add margin to separate legend from plot
    plot.margin = margin(10, 20, 30, 10)  # Increased bottom margin to accommodate rotated labels
  )

# Add caption below the plot
cat("\n**Figure 1B.** Genotype flow visualization across a single family (Family ID: 534). Each horizontal line represents a genomic position on chromosome NW_019211454.1, with colors indicating genotypes (yellow = AA, green = AB, blue = BB). Family members are arranged in generational order, showing the inheritance pattern from parents to offspring. The y-axis shows the genomic position in megabases (Mb).")

# save fam_534_F0AA_F1AA for the Source data excel file
#write.csv(fam_534_F0AA_F1AA,
#          file = file.path(source_dir, "Fig1B_single_family_tracks.csv"),
#          row.names = FALSE)

```

## Figure 1D: Crossing F1 mother AA with F1 father AB, all families

```{r AAxAB-pooled}
# Set the families (include only families with at least one adult F2)
family <- grep("grndat|grnson", gt$ID, value=TRUE) %>%
  str_extract("[^_]+") %>%
  unique()

# Define a list to store all data frames per family
obs <- list()

for (fam in family) {
obs[[fam]] <- table %>%
  dplyr::select(starts_with(fam)) %>%
    # Add site and chromosome information
    rownames_to_column("site") %>%
    mutate(chrom = str_split(site, pattern = "_", simplify = TRUE)[, 2]) %>%
    # Then filter for the genotype patterns we want
    dplyr::filter_at(vars(matches("_fnd")), all_vars(. == "AB")) %>%  # F0 Female must be AB
    dplyr::filter_at(vars(matches("_son")), all_vars(. == "AB")) %>%  # Sons must be AB
    dplyr::filter_at(vars(matches("_dat")), all_vars(. == "AA")) %>%  # Daughters must be AA
    # Select only the grandchildren columns and keep site/chrom info
    dplyr::select(site, chrom, contains("grn")) %>%
    # Reshape the data
    tidyr::pivot_longer(cols = -c(site, chrom)) %>%
    dplyr::rename(Sample = name, gt = value) %>%
    # filter for sites 
    left_join(filtered_sites, by = c("site", "chrom", "Sample", "gt")) %>%
    na.omit() %>% # Count genotypes
    dplyr::count(Sample, gt, chrom, .drop = FALSE) %>%
  dplyr::filter(gt %in% c("AA", "BB", "AB")) %>%
  mutate(n = as.numeric(n)) %>%
    group_by(Sample) %>%
  mutate(total = as.numeric(sum(n))) %>%
  mutate(sex = case_when(
      grepl("son", Sample) ~ "Male",
      grepl("dat", Sample) ~ "Female"
    )) %>%
    # Add family column
    mutate(family = fam)
}

# Bind all families together into a final data frame containing all observed counts
observed <- do.call("rbind", obs) %>% 
  mutate(Sample = as.character(Sample))

# Create samples dataframe
samples <- data.frame(
  Sample = rep(unique(observed$Sample), each = 3), 
  gt = rep(c("AA", "AB", "BB"))
) %>%   
  mutate(family = str_extract(Sample, "^[0-9]+"))

# Join and create final dataset
samples_obs_filtered_AAxAB <- left_join(samples, observed, by = c("family", "Sample", "gt")) %>% 
  mutate(sex = case_when(
    grepl("son", Sample) ~ "Male",
    grepl("dat", Sample) ~ "Female"
  )) %>%
  group_by(Sample, chrom) %>%
    mutate(total = as.numeric(sum(n))) %>%
    dplyr::mutate(prop = n/total) %>%
  na.omit()

### Summary of sites filtered in each sample
# Create a table showing sites filtered per sample
sites_per_sample <- table %>%
  rownames_to_column("site") %>%
  tidyr::pivot_longer(cols = -site, names_to = "Sample", values_to = "gt") %>%
  # Filter out NA values before counting
  filter(!is.na(gt)) %>%
  group_by(Sample) %>%
  summarise(
    total_sites = n(),
    .groups = 'drop'
  )

filtered_sites_per_sample <- filtered_sites %>%
  group_by(Sample) %>%
  summarise(
    filtered_sites = n(),
    .groups = 'drop'
  )

# Join and calculate differences
sites_comparison <- sites_per_sample %>%
  left_join(filtered_sites_per_sample, by = "Sample") %>%
  mutate(
    filtered_sites = replace_na(filtered_sites, 0),
    sites_removed = total_sites - filtered_sites,
    percent_removed = round((sites_removed / total_sites) * 100, 2)
  ) %>%
  arrange(desc(sites_removed))

# Display the table
sites_comparison %>%
  select(Sample, total_sites, filtered_sites, sites_removed, percent_removed) %>%
  rename(
    "Sample" = Sample,
    "Total Sites" = total_sites,
    "Sites Kept" = filtered_sites,
    "Sites Removed" = sites_removed,
    "% Removed" = percent_removed
  ) %>%
  knitr::kable(
    digits = 2,
    caption = "Number of sites filtered per sample",
    format = "html"
  ) %>%
  kableExtra::kable_styling(
    bootstrap_options = "striped",
    full_width = TRUE,
    position = "left",
    fixed_thead = TRUE
  ) %>%
  kableExtra::scroll_box(
    width = "100%", 
    height = "300px"
  )

```

```{r plot-AAxAB-pooled, echo=TRUE, eval=TRUE, message=FALSE}
# Set minimum sites threshold
site_threshold = as.numeric(10)

# Create summary dataframe for labels, summing across all chromosomes (for pulled data only)
summary_df_filtered <- samples_obs_filtered_AAxAB %>%
   filter(total >= site_threshold) %>%
  group_by(sex) %>%  # Remove chromosome grouping
  summarise(
    num_samples = n_distinct(Sample),  # Count unique samples
    total_sites = sum(total, na.rm = TRUE),  # Sum all sites
    .groups = 'drop'
  )

# Create per-chromosome summary for faceted plot
summary_df_filtered_by_chrom <- samples_obs_filtered_AAxAB %>%
  group_by(chrom, sex) %>%
  summarise(
    num_samples = n_distinct(Sample),
    total_sites = sum(total, na.rm = TRUE),
    .groups = 'drop'
  )

# Plot 1: Pulled data with totals across all chromosomes
ggplot(samples_obs_filtered_AAxAB %>% filter(total >= site_threshold), 
       aes(fill = gt, y = prop, x = factor(sex, levels = c("Female", "Male")))) +
  geom_bar(position = "fill", stat = "identity", width = 0.5) +
  ylab("Genotype Proportion") +
  labs(title = "Offspring Genotype, Crossing (AB) Male x (AA) Female
 Filtered sites", 
       fill = "Genotype") +
  theme_classic() +
  theme(
    axis.title.x = element_blank(),
    axis.text.x = element_blank(),
    axis.ticks.x = element_blank(),
    axis.text.y = element_text(size = 12),
    strip.text.x = element_text(size = 12),
    panel.spacing = unit(1, "lines"),
    plot.margin = margin(t = 10, r = 10, b = 60, l = 10)
  ) +
  scale_fill_manual(values = c("#ffbf00", "#66b032", "#1982c4")) +
  # Add sample count labels (only for pulled data)
  geom_text(data = summary_df_filtered,
            aes(x = sex, 
                y = -0.03,
                label = paste0(num_samples, " ", sex, "s")),
            inherit.aes = FALSE,
            size = 2.8,
            vjust = 1) +
  # Add total sites label with comma formatting (only for pulled data)
  geom_text(data = summary_df_filtered,
            aes(x = sex, 
                y = -0.18,
                label = paste0(format(total_sites, big.mark = ","), " sites")),
            inherit.aes = FALSE,
            size = 2.8,
            vjust = 0.5)

 # save samples_obs_filtered_AAxAB for the Source data excel file
#write.csv(samples_obs_filtered_AAxAB,
#          file = file.path(source_dir, "Fig1D_AAxAB_barplot.csv"),
#          row.names = FALSE)

 # save summary_df_filtered for the Source data excel file
#write.csv(summary_df_filtered,
#          file = file.path(source_dir, "Fig1D_sites_per_sample.csv"),
#          row.names = FALSE)
```

```{r clean plot-no lables, echo=TRUE, eval=TRUE, message=FALSE}
# clean plot without labels , for the MS
ggplot(samples_obs_filtered_AAxAB %>% filter(total >= site_threshold), 
       aes(fill = gt, y = prop, x = factor(sex, levels = c("Female", "Male")))) +
  geom_bar(position = "fill", stat = "identity", width = 0.7) +
  ylab("Genotype
Proportion") +
  labs(title = "Offspring Genotype, Crossing (AB) Male x (AA) Female
 Filtered sites", 
       fill = "Genotype") +
  theme_classic() +
  theme(
    axis.title.x = element_blank(),
    axis.text.x = element_blank(),
    axis.ticks.x = element_blank(),
    axis.text.y = element_text(size = 14),
    axis.title.y = element_text(size = 16),
    strip.text.x = element_text(size = 12),
    panel.spacing = unit(1, "lines"),
    plot.margin = margin(t = 10, r = 10, b = 0, l = 10)
  ) +
  scale_fill_manual(values = c("#ffbf00", "#66b032", "#1982c4"))# +
  #facet_wrap(~family)
```

## Figure 2: RNAseq of males and females 

```{r rnaseq-plot, echo=TRUE}
rna_dat <- read_csv("/Users/nuriteliash/Documents/GitHub/Varroa-RNAseq-Zheng-2025/data/rna-seq.csv")

rna_dat_summary <- rna_dat %>%
  group_by(family, sex) %>%
  summarize(
    het_prop = sum(het) / sum(het + homo),  # Percentage of heterozygous
    count = sum(het + homo),                # Total trials
    sem = sqrt((het_prop * (1 - het_prop)) / count),  # Standard error of the mean
    ci_lower = het_prop - qt(0.975, df = count - 1) * sem,  # Lower confidence interval
    ci_upper = het_prop + qt(0.975, df = count - 1) * sem   # Upper confidence interval
  )

# Plot the data
rna_dat_summary %>%
  ggplot(aes(x = sex, y = het_prop, colour = sex)) +
  geom_boxplot() +  
  theme_bw() +
  ylab("Proportion of heterozygous sites") +
  xlab("Sex") +
  theme(legend.position = "bottom") +
 # scale_y_continuous(labels = scales::percent_format())  + 
  scale_color_manual(values = c("Male" = "blue", "Female" = "#E31C79")) + # Custom colors
  theme(legend.position = "none")

cat("\n**Figure 2.** RNA-seq analysis of heterozygous site proportions in male and female mites. Each point represents the mean proportion of heterozygous sites per individual mite, with error bars showing 95% confidence intervals. Points are connected by lines to show family relatedness of mother and son. The y-axis shows the proportion of heterozygous sites as a percentage of total sites analyzed.")

# save rna_dat for the Source data excel file
#write.csv(rna_dat,
#          file = file.path(source_dir, "Fig2_hetprop_RNAseq.csv"),
#          row.names = FALSE)

```

Do  heterozygocity proportion differ significantly between male and female mites?
```{r rnaseq-stats}
# Reshape to wide format
df_wide <- rna_dat_summary %>%
  select(family, sex, het_prop) %>%
  pivot_wider(names_from = sex, values_from = het_prop)

# Paired t-test
t.test(df_wide$Female, df_wide$Male, paired = TRUE)
```
