---
title: "figures-for-varroa-pedigree-MS"
date: "2023-02-20"
output:
  html_document:
    code_folding: hide
    theme: cerulean
    toc: yes
    toc_depth: 5
    toc_float:
      collapsed: yes
      df_print: paged
editor_options: 
  chunk_output_type: console
  markdown: 
    wrap: 72
---

```{=html}
<style type="text/css">
  body{
  font-size: 12pt;
}
</style>
```

### load libraries
```{r include=FALSE}
library("tidyverse")
library("dplyr")
library("ggplot2")
library("ggpubr")
library("scales")
library("ggpubr")
library("gridExtra")
library("grid")
library("GGally")
library("vcfR") # for extracting genotype data from a vcf file
library("janitor")
library("splitstackshape")
library("plotly")
library("kableExtra")
library("pastecs") # for stats stat.desc()

knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                     fig.width = 10,
                      fig.asp = 0.4,
                      out.width = "100%")
```

the VCF was filtered with the following parameters:  
```{r eval=FALSE, echo=TRUE}
vcftools --vcf snponly_freebayes.vcf --chr NW_019211454.1 --chr NW_019211455.1 --chr NW_019211456.1 --chr NW_019211457.1 --chr NW_019211458.1 --chr NW_019211459.1 --chr NW_019211460.1 --max-alleles 2 --minQ 15000 --minDP 16 --maxDP 40 --max-missing 0.5 --maf 0.2 --recode --recode-INFO-all --out Q15000BIALLDP16HDP40mis.5maf.2Chr7  
```

### load VCF file

```{r}
vcf <- read.vcfR("/Users/nuriteliash/Documents/GitHub/varroa-pedigree-study/data/vcf_filter/minQ_filter/Q15000BIALLDP16HDP40mis.5maf.2Chr7.recode.vcf", verbose = FALSE )
vcf
# extract the genotype for each site in each individual
gt <- extract.gt(vcf, element = "GT") 
gt <- as.data.frame(t(gt)) %>%
  rownames_to_column("ID")
#clean the ID names
gt$ID <- sub("_[^_]+$", "", gt$ID)

table <- gt %>%
  t() %>%
  as.data.frame() %>%
  row_to_names(row_number = 1) %>%
  dplyr::select(contains(c("son", "dat", "fnd"))) %>%  # Keep only F0, F1, and F2 adults
  # 🔹 Remove columns containing specific substrings
  dplyr::select(-matches("sis|479-2|565-2|535_1|535_3")) %>%
  # 🔹 Convert GT values from numbers to letters
  mutate(across(everything(), ~ case_when(
    . == "0/0" ~ "AA",
    . == "0/1" ~ "AB",
    . == "1/1" ~ "BB",
    TRUE ~ .  # Keep other values unchanged
  )))
```

# Main figures

## Figure 2A. Proportion of maternal genotype in F2 mites
the original figure can be re-run in here:  
~/Documents/GitHub/varroa-pedigree-study/R_scripts/pedigree-hetero-count-Q15000.Rmd

~/Documents/GitHub/varroa-pedigree-study/R_scripts/After_revision/site_quality_exploration.Rmd

## Figure 2B. F1 male (AA) x Female (AB)
```{r}

```

### Analysis of all sites
Crossing homozygotic male, with heterozygotic Female (5) F1 male (AA) x Female (AB)

```{r all-sites-analysis}
# Set the families (include only families with at least one adult F2)
family <- grep("grndat|grnson", gt$ID, value=TRUE) %>%
  str_extract("[^_]+") %>%
  unique()

# Define a list to store all data frames per family
obs <- list()

for (fam in family) {
  obs[[fam]] <- table %>%
    dplyr::select(starts_with(fam)) %>%
    # Add site and chromosome information
    rownames_to_column("site") %>%
    mutate(chrom = str_split(site, pattern = "_", simplify = TRUE)[, 2]) %>%
    # Filter for the genotype patterns we want
    dplyr::filter_at(vars(matches("_fnd")), all_vars(. == "AA")) %>%  # F0 Female must be AA
    dplyr::filter_at(vars(matches("_son")), all_vars(. == "AA")) %>%  # Sons must be AA
    dplyr::filter_at(vars(matches("_dat")), all_vars(. == "AB")) %>%  # Daughters must be AB
    # Select only the grandchildren columns and keep site/chrom info
    dplyr::select(site, chrom, contains("grn")) %>%
    # Reshape the data
    tidyr::pivot_longer(cols = -c(site, chrom)) %>%
    dplyr::rename(Sample = name, gt = value) %>%
    # Count genotypes
    dplyr::count(Sample, gt, chrom, .drop = FALSE) %>%
    dplyr::filter(gt %in% c("AA", "BB", "AB")) %>%
    mutate(n = as.numeric(n)) %>%
    group_by(Sample) %>%
    mutate(total = as.numeric(sum(n))) %>%
    mutate(sex = case_when(
      grepl("son", Sample) ~ "Male",
      grepl("dat", Sample) ~ "Female"
    )) %>%
    # Add family column
    mutate(family = fam)
}

# Bind all families together into a final data frame containing all observed counts
observed <- do.call("rbind", obs) %>% 
  mutate(Sample = as.character(Sample))

# Create samples dataframe
samples <- data.frame(
  Sample = rep(unique(observed$Sample), each = 3), 
  gt = rep(c("AA", "AB", "BB"))
) %>%   
  mutate(family = str_extract(Sample, "^[0-9]+"))

# Join and create final dataset
samples_obs_all <- left_join(samples, observed, by = c("family", "Sample", "gt")) %>% 
  mutate(sex = case_when(
    grepl("son", Sample) ~ "Male",
    grepl("dat", Sample) ~ "Female"
  )) %>%
  group_by(Sample, chrom) %>%
  mutate(total = as.numeric(sum(n))) %>%
  dplyr::mutate(prop = n/total) %>%
  na.omit()
```

now we count only the filtered sites, based on VAF and QAF values     

rule:       
  group_by(family, site, Sample) %>%  # Keep sites grouped by family    
  filter(   
    (GT == "AA" & QAF < 0.03 & VAF < 0.03) |   # Homozygous Reference (AA)    
    (GT == "AB" & QAF > 0.4 & QAF < 0.6 & VAF > 0.4 & VAF < 0.6) |  # Heterozygous (AB)   
    (GT == "BB" & QAF > 0.8 & VAF > 0.8)    # Homozygous Alternate (BB))     

### Analysis of filtered sites (site quality QAF and VAF)
The analysis was done in "~/Documents/GitHub/varroa-pedigree-study/R_scripts/After_revision/site_quality_exploration.Rmd"

```{r filtered-sites-analysis}
# Load the pre-filtered sites per family and sample
filtered_sites <- read_csv("/Users/nuriteliash/Documents/GitHub/varroa-pedigree-study/data/filtered_sites.csv") %>%
    mutate(family = as.character(family),
           Sample = as.character(Sample),
           # Extract Chromosome from the site column
           chrom = str_split(site, pattern = "_", simplify = TRUE)[, 2]) %>% rename(gt = GT)

# Set the families (include only families with at least one adult F2)
family <- grep("grndat|grnson", gt$ID, value=TRUE) %>%
  str_extract("[^_]+") %>%
  unique()

# Define a list to store all data frames per family
obs <- list()

for (fam in family) {
  obs[[fam]] <- table %>%
    dplyr::select(starts_with(fam)) %>%
    # Add site and chromosome information
    rownames_to_column("site") %>%
    mutate(chrom = str_split(site, pattern = "_", simplify = TRUE)[, 2]) %>%
    # Then filter for the genotype patterns we want
    dplyr::filter_at(vars(matches("_fnd")), all_vars(. == "AA")) %>%  # F0 Female must be AA
    dplyr::filter_at(vars(matches("_son")), all_vars(. == "AA")) %>%  # Sons must be AA
    dplyr::filter_at(vars(matches("_dat")), all_vars(. == "AB")) %>%  # Daughters must be AB
    # Select only the grandchildren columns and keep site/chrom info
    dplyr::select(site, chrom, contains("grn")) %>%
    # Reshape the data
    tidyr::pivot_longer(cols = -c(site, chrom)) %>%
    dplyr::rename(Sample = name, gt = value) %>%
    # filter for sites 
     left_join(filtered_sites, by = c("site", "chrom", "Sample",  "gt")) %>% #view()
    na.omit() %>% # Count genotypes
    dplyr::count(Sample, gt, chrom, .drop = FALSE) %>%
    dplyr::filter(gt %in% c("AA", "BB", "AB")) %>%
    mutate(n = as.numeric(n)) %>%
    group_by(Sample) %>%
    mutate(total = as.numeric(sum(n))) %>%
    mutate(sex = case_when(
      grepl("son", Sample) ~ "Male",
      grepl("dat", Sample) ~ "Female"
    )) %>%
    # Add family column
    mutate(family = fam)
}

# Bind all families together into a final data frame containing all observed counts
observed <- do.call("rbind", obs) %>% 
  mutate(Sample = as.character(Sample))

# Create samples dataframe
Sample <- data.frame(
  Sample = rep(unique(observed$Sample), each = 3), 
  gt = rep(c("AA", "AB", "BB"))
) %>%   
  mutate(family = str_extract(Sample, "^[0-9]+"))

# Join and create final dataset
samples_obs_filtered <- left_join(Sample, observed, by = c("family", "Sample", "gt")) %>% 
  mutate(sex = case_when(
    grepl("son", Sample) ~ "Male",
    grepl("dat", Sample) ~ "Female"
  )) %>%
  group_by(Sample, chrom) %>%
  mutate(total = as.numeric(sum(n))) %>%
  dplyr::mutate(prop = n/total) %>%
  na.omit()
```

### Comparison of all sites vs filtered sites
```{r compare-datasets}
# Create a table showing sites filtered per sample
cat("\nSites filtered per sample:\n")
sites_per_sample <- table %>%
  rownames_to_column("site") %>%
  tidyr::pivot_longer(cols = -site, names_to = "Sample", values_to = "gt") %>%
  # Filter out NA values before counting
  filter(!is.na(gt)) %>%
  group_by(Sample) %>%
  summarise(
    total_sites = n(),
    .groups = 'drop'
  )

filtered_sites_per_sample <- filtered_sites %>%
  group_by(Sample) %>%
  summarise(
    filtered_sites = n(),
    .groups = 'drop'
  )

# Join and calculate differences
sites_comparison <- sites_per_sample %>%
  left_join(filtered_sites_per_sample, by = "Sample") %>%
  mutate(
    filtered_sites = replace_na(filtered_sites, 0),
    sites_removed = total_sites - filtered_sites,
    percent_removed = round((sites_removed / total_sites) * 100, 2)
  ) %>%
  arrange(desc(sites_removed))

# Display the table
knitr::kable(sites_comparison %>%
  select(Sample, total_sites, filtered_sites, sites_removed, percent_removed) %>%
  rename(
    "Sample" = Sample,
    "Total Sites" = total_sites,
    "Sites Kept" = filtered_sites,
    "Sites Removed" = sites_removed,
    "% Removed" = percent_removed
  ),
  digits = 2,
  caption = "Number of sites filtered per sample") %>%
  kable_styling(full_width = TRUE)

```

### Filtered for QAF and VAF, minimum sites per sample = 10
```{r comparing-all-and-filtered-data-sites_min10}
site_threshold = as.numeric(10)

## all sites
# First, create the summary dataframe for labels
summary_df_all <- samples_obs_all %>%
  group_by(chrom, sex) %>%
  summarise(
    num_samples = length(unique(Sample)),
    total_sites = sum(total, na.rm = TRUE),
    .groups = 'drop'
  )

# plot pulled data
ggplot(samples_obs_all %>% filter(total >= site_threshold), 
       aes(fill = gt, y = prop, x = factor(sex, levels = c("Female", "Male")))) +
  geom_bar(position = "fill", stat = "identity", width = 0.5) +  # Made bars narrower (changed from 0.7 to 0.5)
  #facet_wrap(~chrom, nrow = 1) +
  ylab("Genotype Proportion") +
  labs(title = "Offspring Genotype, Crossing (AA) Male x (AB) Female
All sites, sites per sample => 10", 
       fill = "Genotype") +
  theme_classic() +
  theme(
    axis.title.x = element_blank(),
    axis.text.x = element_blank(),
    axis.ticks.x = element_blank(),
    axis.text.y = element_text(size = 12),
    strip.text.x = element_text(size = 12),
    panel.spacing = unit(1, "lines"),  # Reduced panel spacing (changed from 2 to 1)
    plot.margin = margin(t = 10, r = 10, b = 60, l = 10)
  ) +
  scale_fill_manual(values = c("#ffbf00", "#66b032", "#1982c4")) +
  # Add sample count labels
  geom_text(data = summary_df_all,
            aes(x = sex, 
                y = -0.03,
                label = paste0(num_samples, " ", sex, "s")),
            inherit.aes = FALSE,
            size = 2.8,
            vjust = 1) +
  # Add sites count labels with comma formatting
  geom_text(data = summary_df_all,
            aes(x = sex, 
                y = -0.18,
                label = paste0(format(total_sites, big.mark = ","), " sites")),
            inherit.aes = FALSE,
            size = 2.8,
            vjust = 0.5)

# plot facet data
ggplot(samples_obs_all %>% filter(total >= site_threshold), 
       aes(fill = gt, y = prop, x = factor(sex, levels = c("Female", "Male")))) +
  geom_bar(position = "fill", stat = "identity", width = 0.5) +  # Made bars narrower (changed from 0.7 to 0.5)
  facet_wrap(~chrom, nrow = 1) +
  ylab("Genotype Proportion") +
  labs(title = "Offspring Genotype, Crossing (AA) Male x (AB) Female
All sites, sites per sample => 10", 
       fill = "Genotype") +
  theme_classic() +
  theme(
    axis.title.x = element_blank(),
    axis.text.x = element_blank(),
    axis.ticks.x = element_blank(),
    axis.text.y = element_text(size = 12),
    strip.text.x = element_text(size = 12),
    panel.spacing = unit(1, "lines"),  # Reduced panel spacing (changed from 2 to 1)
    plot.margin = margin(t = 10, r = 10, b = 60, l = 10)
  ) +
  scale_fill_manual(values = c("#ffbf00", "#66b032", "#1982c4")) +
  # Add sample count labels
  geom_text(data = summary_df_all,
            aes(x = sex, 
                y = -0.03,
                label = paste0(num_samples, " ", sex, "s")),
            inherit.aes = FALSE,
            size = 2.8,
            vjust = 1) +
  # Add sites count labels with comma formatting
  geom_text(data = summary_df_all,
            aes(x = sex, 
                y = -0.18,
                label = paste0(format(total_sites, big.mark = ","), " sites")),
            inherit.aes = FALSE,
            size = 2.8,
            vjust = 0.5)

## filtered sites

# First, create the summary dataframe for labels
summary_df_filtered <- samples_obs_filtered %>%
  group_by(chrom, sex) %>%
  summarise(
    num_samples = length(unique(Sample)),
    total_sites = sum(total, na.rm = TRUE),
    .groups = 'drop'
  )

# plot pulled data 
ggplot(samples_obs_filtered %>% filter(total >= site_threshold), 
       aes(fill = gt, y = prop, x = factor(sex, levels = c("Female", "Male")))) +
  geom_bar(position = "fill", stat = "identity", width = 0.5) +  # Made bars narrower (changed from 0.7 to 0.5)
 # facet_wrap(~chrom, nrow = 1) +
  ylab("Genotype Proportion") +
  labs(title = "Offspring Genotype, Crossing (AA) Male x (AB) Female
 Filtered sites, sites per sample => 10", 
       fill = "Genotype") +
  theme_classic() +
  theme(
    axis.title.x = element_blank(),
    axis.text.x = element_blank(),
    axis.ticks.x = element_blank(),
    axis.text.y = element_text(size = 12),
    strip.text.x = element_text(size = 12),
    panel.spacing = unit(1, "lines"),  # Reduced panel spacing (changed from 2 to 1)
    plot.margin = margin(t = 10, r = 10, b = 60, l = 10)
  ) +
  scale_fill_manual(values = c("#ffbf00", "#66b032", "#1982c4")) +
  # Add sample count labels
  geom_text(data = summary_df_filtered,
            aes(x = sex, 
                y = -0.03,
                label = paste0(num_samples, " ", sex, "s")),
            inherit.aes = FALSE,
            size = 2.8,
            vjust = 1) +
  # Add sites count labels with comma formatting
  geom_text(data = summary_df_filtered,
            aes(x = sex, 
                y = -0.18,
                label = paste0(format(total_sites, big.mark = ","), " sites")),
            inherit.aes = FALSE,
            size = 2.8,
            vjust = 0.5)

# plot facet data 
ggplot(samples_obs_filtered %>% filter(total >= site_threshold), 
       aes(fill = gt, y = prop, x = factor(sex, levels = c("Female", "Male")))) +
  geom_bar(position = "fill", stat = "identity", width = 0.5) +  # Made bars narrower (changed from 0.7 to 0.5)
  facet_wrap(~chrom, nrow = 1) +
  ylab("Genotype Proportion") +
  labs(title = "Offspring Genotype, Crossing (AA) Male x (AB) Female
 Filtered sites, sites per sample => 10", 
       fill = "Genotype") +
  theme_classic() +
  theme(
    axis.title.x = element_blank(),
    axis.text.x = element_blank(),
    axis.ticks.x = element_blank(),
    axis.text.y = element_text(size = 12),
    strip.text.x = element_text(size = 12),
    panel.spacing = unit(1, "lines"),  # Reduced panel spacing (changed from 2 to 1)
    plot.margin = margin(t = 10, r = 10, b = 60, l = 10)
  ) +
  scale_fill_manual(values = c("#ffbf00", "#66b032", "#1982c4")) +
  # Add sample count labels
  geom_text(data = summary_df_filtered,
            aes(x = sex, 
                y = -0.03,
                label = paste0(num_samples, " ", sex, "s")),
            inherit.aes = FALSE,
            size = 2.8,
            vjust = 1) +
  # Add sites count labels with comma formatting
  geom_text(data = summary_df_filtered,
            aes(x = sex, 
                y = -0.18,
                label = paste0(format(total_sites, big.mark = ","), " sites")),
            inherit.aes = FALSE,
            size = 2.8,
            vjust = 0.5)
## Box plots
# All sites box plot
ggplot(samples_obs_all %>% 
       filter(total >= site_threshold) %>%
       filter(gt == "AA"), 
       aes(x = factor(sex, levels = c("Female", "Male")), y = prop, color = sex)) +
  geom_boxplot(fill = "grey80", outlier.shape = NA, color = "black") +
  geom_point(position = position_jitter(width = 0.2), size = 2, alpha = 0.7) +
  facet_wrap(~chrom, nrow = 2) +
  ylab("Proportion of AA Genotype") +
  xlab("Sex") +
  labs(title = "Distribution of AA Genotype Proportion per Chromosome
All sites, sites per sample => 10") +
  theme_classic() +
  theme(
    strip.background = element_rect(fill = "white", color = "black"),
    axis.text = element_text(size = 12),
    strip.text = element_text(size = 12),
    panel.spacing = unit(1, "lines"),
    plot.margin = margin(t = 10, r = 10, b = 10, l = 10)
  ) +
  scale_color_manual(values = c("Female" = "red", "Male" = "blue")) +
  coord_cartesian(ylim = c(0, 1))

# Filtered sites box plot
ggplot(samples_obs_filtered %>% 
       filter(total >= site_threshold) %>%
       filter(gt == "AA"), 
       aes(x = factor(sex, levels = c("Female", "Male")), y = prop, color = sex)) +
  geom_boxplot(fill = "grey80", outlier.shape = NA, color = "black") +
  geom_point(position = position_jitter(width = 0.2), size = 2, alpha = 0.7) +
  facet_wrap(~chrom, nrow = 2) +
  ylab("Proportion of AA Genotype") +
  xlab("Sex") +
  labs(title = "Distribution of AA Genotype Proportion per Chromosome
Filtered sites, sites per sample => 10") +
  theme_classic() +
  theme(
    strip.background = element_rect(fill = "white", color = "black"),
    axis.text = element_text(size = 12),
    strip.text = element_text(size = 12),
    panel.spacing = unit(1, "lines"),
    plot.margin = margin(t = 10, r = 10, b = 10, l = 10)
  ) +
  scale_color_manual(values = c("Female" = "red", "Male" = "blue")) +
  coord_cartesian(ylim = c(0, 1))
```


### Filtered for QAF and VAF, minimum sites per sample = 40
```{r comparing-all-and-filtered-data-sites_min40}
site_threshold = as.numeric(40)

## all sites
# First, create the summary dataframe for labels
summary_df_all <- samples_obs_all %>%
  group_by(chrom, sex) %>%
  summarise(
    num_samples = length(unique(Sample)),
    total_sites = sum(total, na.rm = TRUE),
    .groups = 'drop'
  )

# plot pulled data
ggplot(samples_obs_all %>% filter(total >= site_threshold), 
       aes(fill = gt, y = prop, x = factor(sex, levels = c("Female", "Male")))) +
  geom_bar(position = "fill", stat = "identity", width = 0.5) +  # Made bars narrower (changed from 0.7 to 0.5)
  #facet_wrap(~chrom, nrow = 1) +
  ylab("Genotype Proportion") +
  labs(title = "Offspring Genotype, Crossing (AA) Male x (AB) Female
All sites, sites per sample => 40", 
       fill = "Genotype") +
  theme_classic() +
  theme(
    axis.title.x = element_blank(),
    axis.text.x = element_blank(),
    axis.ticks.x = element_blank(),
    axis.text.y = element_text(size = 12),
    strip.text.x = element_text(size = 12),
    panel.spacing = unit(1, "lines"),  # Reduced panel spacing (changed from 2 to 1)
    plot.margin = margin(t = 10, r = 10, b = 60, l = 10)
  ) +
  scale_fill_manual(values = c("#ffbf00", "#66b032", "#1982c4")) +
  # Add sample count labels
  geom_text(data = summary_df_all,
            aes(x = sex, 
                y = -0.03,
                label = paste0(num_samples, " ", sex, "s")),
            inherit.aes = FALSE,
            size = 2.8,
            vjust = 1) +
  # Add sites count labels with comma formatting
  geom_text(data = summary_df_all,
            aes(x = sex, 
                y = -0.18,
                label = paste0(format(total_sites, big.mark = ","), " sites")),
            inherit.aes = FALSE,
            size = 2.8,
            vjust = 0.5)

# plot facet data
ggplot(samples_obs_all %>% filter(total >= site_threshold), 
       aes(fill = gt, y = prop, x = factor(sex, levels = c("Female", "Male")))) +
  geom_bar(position = "fill", stat = "identity", width = 0.5) +  # Made bars narrower (changed from 0.7 to 0.5)
  facet_wrap(~chrom, nrow = 1) +
  ylab("Genotype Proportion") +
  labs(title = "Offspring Genotype, Crossing (AA) Male x (AB) Female
All sites, sites per sample => 40", 
       fill = "Genotype") +
  theme_classic() +
  theme(
    axis.title.x = element_blank(),
    axis.text.x = element_blank(),
    axis.ticks.x = element_blank(),
    axis.text.y = element_text(size = 12),
    strip.text.x = element_text(size = 12),
    panel.spacing = unit(1, "lines"),  # Reduced panel spacing (changed from 2 to 1)
    plot.margin = margin(t = 10, r = 10, b = 60, l = 10)
  ) +
  scale_fill_manual(values = c("#ffbf00", "#66b032", "#1982c4")) +
  # Add sample count labels
  geom_text(data = summary_df_all,
            aes(x = sex, 
                y = -0.03,
                label = paste0(num_samples, " ", sex, "s")),
            inherit.aes = FALSE,
            size = 2.8,
            vjust = 1) +
  # Add sites count labels with comma formatting
  geom_text(data = summary_df_all,
            aes(x = sex, 
                y = -0.18,
                label = paste0(format(total_sites, big.mark = ","), " sites")),
            inherit.aes = FALSE,
            size = 2.8,
            vjust = 0.5)

## filtered sites

# First, create the summary dataframe for labels
summary_df_filtered <- samples_obs_filtered %>%
  group_by(chrom, sex) %>%
  summarise(
    num_samples = length(unique(Sample)),
    total_sites = sum(total, na.rm = TRUE),
    .groups = 'drop'
  )

# plot pulled data 
ggplot(samples_obs_filtered %>% filter(total >= site_threshold), 
       aes(fill = gt, y = prop, x = factor(sex, levels = c("Female", "Male")))) +
  geom_bar(position = "fill", stat = "identity", width = 0.5) +  # Made bars narrower (changed from 0.7 to 0.5)
 # facet_wrap(~chrom, nrow = 1) +
  ylab("Genotype Proportion") +
  labs(title = "Offspring Genotype, Crossing (AA) Male x (AB) Female
 Filtered sites, sites per sample => 40", 
       fill = "Genotype") +
  theme_classic() +
  theme(
    axis.title.x = element_blank(),
    axis.text.x = element_blank(),
    axis.ticks.x = element_blank(),
    axis.text.y = element_text(size = 12),
    strip.text.x = element_text(size = 12),
    panel.spacing = unit(1, "lines"),  # Reduced panel spacing (changed from 2 to 1)
    plot.margin = margin(t = 10, r = 10, b = 60, l = 10)
  ) +
  scale_fill_manual(values = c("#ffbf00", "#66b032", "#1982c4")) +
  # Add sample count labels
  geom_text(data = summary_df_filtered,
            aes(x = sex, 
                y = -0.03,
                label = paste0(num_samples, " ", sex, "s")),
            inherit.aes = FALSE,
            size = 2.8,
            vjust = 1) +
  # Add sites count labels with comma formatting
  geom_text(data = summary_df_filtered,
            aes(x = sex, 
                y = -0.18,
                label = paste0(format(total_sites, big.mark = ","), " sites")),
            inherit.aes = FALSE,
            size = 2.8,
            vjust = 0.5)

# plot facet data 
ggplot(samples_obs_filtered %>% filter(total >= site_threshold), 
       aes(fill = gt, y = prop, x = factor(sex, levels = c("Female", "Male")))) +
  geom_bar(position = "fill", stat = "identity", width = 0.5) +  # Made bars narrower (changed from 0.7 to 0.5)
  facet_wrap(~chrom, nrow = 1) +
  ylab("Genotype Proportion") +
  labs(title = "Offspring Genotype, Crossing (AA) Male x (AB) Female
 Filtered sites, sites per sample => 40", 
       fill = "Genotype") +
  theme_classic() +
  theme(
    axis.title.x = element_blank(),
    axis.text.x = element_blank(),
    axis.ticks.x = element_blank(),
    axis.text.y = element_text(size = 12),
    strip.text.x = element_text(size = 12),
    panel.spacing = unit(1, "lines"),  # Reduced panel spacing (changed from 2 to 1)
    plot.margin = margin(t = 10, r = 10, b = 60, l = 10)
  ) +
  scale_fill_manual(values = c("#ffbf00", "#66b032", "#1982c4")) +
  # Add sample count labels
  geom_text(data = summary_df_filtered,
            aes(x = sex, 
                y = -0.03,
                label = paste0(num_samples, " ", sex, "s")),
            inherit.aes = FALSE,
            size = 2.8,
            vjust = 1) +
  # Add sites count labels with comma formatting
  geom_text(data = summary_df_filtered,
            aes(x = sex, 
                y = -0.18,
                label = paste0(format(total_sites, big.mark = ","), " sites")),
            inherit.aes = FALSE,
            size = 2.8,
            vjust = 0.5)
## Box plots
# All sites box plot
ggplot(samples_obs_all %>% 
       filter(total >= site_threshold) %>%
       filter(gt == "AA"), 
       aes(x = factor(sex, levels = c("Female", "Male")), y = prop, color = sex)) +
  geom_boxplot(fill = "grey80", outlier.shape = NA, color = "black") +
  geom_point(position = position_jitter(width = 0.2), size = 2, alpha = 0.7) +
  facet_wrap(~chrom, nrow = 2) +
  ylab("Proportion of AA Genotype") +
  xlab("Sex") +
  labs(title = "Distribution of AA Genotype Proportion per Chromosome
All sites, sites per sample => 40") +
  theme_classic() +
  theme(
    strip.background = element_rect(fill = "white", color = "black"),
    axis.text = element_text(size = 12),
    strip.text = element_text(size = 12),
    panel.spacing = unit(1, "lines"),
    plot.margin = margin(t = 10, r = 10, b = 10, l = 10)
  ) +
  scale_color_manual(values = c("Female" = "red", "Male" = "blue")) +
  coord_cartesian(ylim = c(0, 1))

# Filtered sites box plot
ggplot(samples_obs_filtered %>% 
       filter(total >= site_threshold) %>%
       filter(gt == "AA"), 
       aes(x = factor(sex, levels = c("Female", "Male")), y = prop, color = sex)) +
  geom_boxplot(fill = "grey80", outlier.shape = NA, color = "black") +
  geom_point(position = position_jitter(width = 0.2), size = 2, alpha = 0.7) +
  facet_wrap(~chrom, nrow = 2) +
  ylab("Proportion of AA Genotype") +
  xlab("Sex") +
  labs(title = "Distribution of AA Genotype Proportion per Chromosome
Filtered sites, sites per sample => 40") +
  theme_classic() +
  theme(
    strip.background = element_rect(fill = "white", color = "black"),
    axis.text = element_text(size = 12),
    strip.text = element_text(size = 12),
    panel.spacing = unit(1, "lines"),
    plot.margin = margin(t = 10, r = 10, b = 10, l = 10)
  ) +
  scale_color_manual(values = c("Female" = "red", "Male" = "blue")) +
  coord_cartesian(ylim = c(0, 1))
```

---

# Archived Code for Future Reference
*The following code sections are preserved for future analysis but are not currently active in the main analysis.*

```{r filtered-sites-analysis-archived}
# Load the pre-filtered sites per family and sample
filtered_sites <- read_csv("/Users/nuriteliash/Documents/GitHub/varroa-pedigree-study/data/filtered_sites.csv") %>%
    mutate(family = as.character(family),
           Sample = as.character(Sample),
           # Extract Chromosome from the site column
           chrom = str_split(site, pattern = "_", simplify = TRUE)[, 2]) %>% rename(gt = GT)

# Define a list to store all data frames per family
obs_count <- list()

for (fam in family) {
  obs_count[[fam]] <- table %>%
    dplyr::select(starts_with(fam)) %>%
    # Add site and chromosome information
    rownames_to_column("site") %>%
    mutate(chrom = str_split(site, pattern = "_", simplify = TRUE)[, 2]) %>%
    # Filter for the genotype patterns we want
    dplyr::filter_at(vars(matches("_fnd")), all_vars(. == "AA")) %>%  # F0 Female must be AA
    dplyr::filter_at(vars(matches("_son")), all_vars(. == "AA")) %>%  # Sons must be AA
    dplyr::filter_at(vars(matches("_dat")), all_vars(. == "AB")) %>%  # Daughters must be AB
    # Select only the grandchildren columns and keep site/chrom info
    dplyr::select(site, chrom, contains("grn")) %>%
    # Reshape the data
    tidyr::pivot_longer(cols = -c(site, chrom)) %>%
    dplyr::rename(Sample = name, gt = value) %>% 
   na.omit() %>% 
     mutate(sex = case_when(
      grepl("son", Sample) ~ "Male",
      grepl("dat", Sample) ~ "Female"
    )) %>%
    # Add family column
    mutate(family = fam) %>% #view()
    left_join(filtered_sites, by = c("site", "chrom", "Sample", "family", "gt")) %>% #view()
    na.omit() # %>% view()
  }

# Bind all families together into a final data frame containing all observed counts
observed_count <- do.call("rbind", obs_count) %>% 
  mutate(Sample = as.character(Sample))
```


## F1 male (AB) x Female (AB)
```{r eval=FALSE, include=FALSE}
# define a list to put all the data frames in
obs <- list()

for (fam in family) {
  
obs[[fam]] <- table %>%
  dplyr::select(starts_with(fam)) %>%
   # dplyr::filter_at(vars(matches("_fnd")), all_vars(. == "AB")) %>% # force F0 Female to be hetero, like her son
  dplyr::filter_at(vars(matches("_son")), all_vars(. == "AB")) %>%
  dplyr::filter_at(vars(matches("_dat")), all_vars(. == "AB")) %>% 
  dplyr::select(contains("grn")) %>%
  tidyr::pivot_longer(everything())  %>% 
  dplyr::rename(sample = name, gt = value) %>%
  dplyr::count(sample, gt, .drop = FALSE) %>% 
  dplyr::filter(gt %in% c("AA", "BB", "AB")) %>%
  mutate(n = as.numeric(n)) %>%
  group_by(sample) %>%
  mutate(total = as.numeric(sum(n))) %>%
  mutate(sex = case_when(
    grepl("son", sample) ~ "Male",
    grepl("dat", sample) ~ "Female"))
}

# bind all families together, to a final data frame containing all observed counts
observed <- do.call("rbind", obs) %>% mutate(sample = as.character(sample))
samples <- data.frame(sample = rep(unique(observed$sample), each =3), gt = rep(c("AA", "AB", "BB")))
   
# define the group of abnormal males based on the overall analsys: (~/Documents/GitHub/varroa-pedigree-study/R_scripts/pedigree-hetero-count-Q15000.Rmd)
abnorm_males = tibble(sample = c("412_413a_grnson",  "400_401a_grnson", "458_459a_grnson", "240_241c_grnson", "426_427b_grnson",  "300_301a_grnson"), sex = "Male", normality = "abnormal")

# for this specific cross, there are no abnormal males: matrnal prop<0.75

samples_obs <- left_join(samples, observed, by=c("sample","gt")) %>% mutate(sex = case_when(
    grepl("son", sample) ~ "Male",
    grepl("dat", sample) ~ "Female")) %>%
    group_by(sample) %>%
    replace(is.na(.), 0) %>%
    mutate(total = as.numeric(sum(n))) %>%
    dplyr::mutate(prop = n/total) %>%
    left_join(abnorm_males, by = "sample") %>% 
    dplyr::select(-sex.y) %>% 
    replace(is.na(.), "normal") %>%
    dplyr::rename(sex = sex.x) %>%
    unite("type", sex,normality, remove = FALSE)

# make a table with the expected proportions for the different modes of reproduction:
Haploid = data.frame(mode = "Haploid", gt = c("AA", "AB", "BB"),prop = c(0.5, 0, 0.5))
Automixis = data.frame(mode = "Automixis", gt = c("AA", "AB", "BB"),prop = c(0, 0.5, 0))
Sexual = data.frame(mode = "Sexual",gt = c("AA", "AB", "BB"),prop = c(0.25, 0.5, 0.25))

modes = rbind(Haploid,Automixis,Sexual)
```

## F1 male (AB) x Female (AA)
```{r eval=FALSE, include=FALSE}
# define a list to put all the data frames in
obs <- list()

for (fam in family) {
  
obs[[fam]] <- table %>%
  dplyr::select(starts_with(fam)) %>%
     dplyr::filter_at(vars(matches("_fnd")), all_vars(. == "AB")) %>% # force F0 Female to be homo, like her son
  dplyr::filter_at(vars(matches("_son")), all_vars(. == "AB")) %>%
  dplyr::filter_at(vars(matches("_dat")), all_vars(. == "AA")) %>% 
  dplyr::select(contains("grn")) %>%
  tidyr::pivot_longer(everything())  %>% 
  dplyr::rename(sample = name, gt = value) %>%
  dplyr::count(sample, gt, .drop = FALSE) %>% 
  dplyr::filter(gt %in% c("AA", "BB", "AB")) %>%
  mutate(n = as.numeric(n)) %>%
  group_by(sample) %>%
  mutate(total = as.numeric(sum(n))) %>%
  mutate(sex = case_when(
    grepl("son", sample) ~ "Male",
    grepl("dat", sample) ~ "Female"))
}

# bind all families together, to a final data frame containing all observed counts
observed <- do.call("rbind", obs) %>% mutate(sample = as.character(sample))
samples <- data.frame(sample = rep(unique(observed$sample), each =3), gt = rep(c("AA", "AB", "BB")))
   
# define the group of abnormal males based on the overall analsys: (~/Documents/GitHub/varroa-pedigree-study/R_scripts/pedigree-hetero-count-Q15000.Rmd)
abnorm_males = tibble(sample = c("412_413a_grnson",  "400_401a_grnson", "458_459a_grnson", "240_241c_grnson", "426_427b_grnson",  "300_301a_grnson"), sex = "Male", normality = "abnormal")

# for this specific cross, there are no abnormal males: matrnal prop<0.75

samples_obs <- left_join(samples, observed, by=c("sample","gt")) %>% mutate(sex = case_when(
    grepl("son", sample) ~ "Male",
    grepl("dat", sample) ~ "Female")) %>%
    group_by(sample) %>%
    replace(is.na(.), 0) %>%
    mutate(total = as.numeric(sum(n))) %>%
    dplyr::mutate(prop = n/total) %>%
    left_join(abnorm_males, by = "sample") %>% 
    dplyr::select(-sex.y) %>% 
    replace(is.na(.), "normal") %>%
    dplyr::rename(sex = sex.x) %>%
    unite("type", sex,normality, remove = FALSE)

# make a table with the expected proportions for the different modes of reproduction:
Haploid = data.frame(mode = "Haploid", gt = c("AA", "AB", "BB"),prop = c(1, 0, 0))
Automixis = data.frame(mode = "Automixis", gt = c("AA", "AB", "BB"),prop = c(1, 0, 0))
Sexual = data.frame(mode = "Sexual",gt = c("AA", "AB", "BB"),prop = c(0.5, 0.5, 0))

modes = rbind(Haploid,Automixis,Sexual)
```

## Heterozygosity Analysis
```{r eval=FALSE, include=FALSE}
### varroa mite DNA
ind_het <- read_delim("/Users/nuriteliash/Documents/GitHub/varroa-pedigree-study/data/vcf_filter/Q40BIALLDP16HDP40mis.5Chr7/Q40BIALLDP16HDP40mis.5Chr7.het", delim = "\t",
           col_names = c("ind","homo_ob", "homo_ex", "nsites", "f"), skip = 1)

# find all 'Female' and 'male' 
Male<- grep("son",ind_het$ind)
Female <- grep("dat|fn|sis",ind_het$ind)

ind_het_sex <- ind_het %>%
  mutate(sex = ifelse(row_number() %in% Female, "Female", ifelse(row_number() %in% Male, "Male", "not-determined"))) %>%
  mutate(hom_prop =  homo_ob/nsites) %>%
  mutate(het_prop = (nsites-homo_ob)/nsites) %>%
 #keep only adult mites, for which sex is absolutly determined (exclude nymphs and eggs ("not determined"))
  dplyr::filter(sex %in% c("Male", "Female"))

# is there a significant difference in the proportion of heterozygotic sites between males and Females?5
wil_var <- wilcox.test(het_prop ~ sex, alternative = "two.sided", data = ind_het_sex)
t_test <- t.test(asin(sqrt(het_prop)) ~ sex, alternative = "two.sided", data = ind_het_sex)
# no significant different (both wilcoxone and welch-test)

# plot heterozygosity proportion, in each sex
p_var_DNA <- ggplot(ind_het_sex) +
    geom_boxplot(aes(x = sex, y = het_prop, fill = sex)) + 
    scale_y_continuous(expand=c(0,0), limits = c(0, 1)) +
    theme_classic() +
    labs(title = "A. Varroa DNA") +
  ylab("Proportion of heterozygotic sites") +
  theme(axis.title.x = element_blank(),
        axis.text.x = element_blank(),
          axis.title.y = element_blank(),
        legend.position = "none",
        legend.title = element_blank(), 
        axis.ticks.x=element_blank()) +scale_fill_manual(values=c("white","black")) 

# get the median and avarage values for the proportions
count_varroa_DNA = ind_het_sex %>%
  group_by(sex) %>%
  summarise(median = median(het_prop), mean = mean(het_prop), n = n(), sd = sd(het_prop)) %>%
  mutate(organism = "varroa") %>%
  mutate(data = "DNA") %>%
  mutate(pvalue = format(round(wil_var$p.value, digits=3)))
```

## Ploidy Analysis
```{r eval=FALSE, include=FALSE}
### Flow-cytometry
data <- read.csv("/Users/nuriteliash/Documents/GitHub/varroa_ploidy/data/ploidy.csv") %>%
  dplyr::mutate(Family = as.character(Family)) %>%
  dplyr::mutate(stage_mature = case_when(
    grepl("adult", Stage) ~ "Mature",
    !grepl("adult", Stage) ~ "Imature")) 

# count the mites in each stage and sex
ploidy_summary = data %>% 
  dplyr::filter(body.part == "Body") %>%
    group_by(Sex, stage_mature) %>% 
     mutate(count_sample = n_distinct(record_number)) %>% #count each smaple once
select(c(Sex,stage_mature,count_sample)) %>% 
    unique()

ploidy_summary

# change the labales 
my_strip_labels <- as_labeller(c(
  "Imature" = "Imature mites
n = 37",
  "Mature" = "Mature mites
n = 38"))

# histogram plot:
p_ploidy_hist = data %>% dplyr::filter(body.part == "Body") %>%
 ggplot(aes(x=Ploidy, color = Sex, fill = Sex)) + 
   geom_histogram(position="dodge") +
    geom_density(alpha=.2) +
   ylab("Count") +
theme_classic2() +
     theme(legend.position = "none") +
ggtitle("Varroa mite ploidy count
Flow-cytometry") +
   facet_grid(.~stage_mature, space = 'free',drop = F,
              labeller = my_strip_labels) # add labels 

# density plot:
p_ploidy_dens = data %>% dplyr::filter(body.part == "Body") %>%
 ggplot(aes(x=Ploidy, color = Sex, fill = Sex)) + 
  # geom_histogram(position="dodge") +
    geom_density(alpha=.2) +
     ylab("Density") +
theme_classic2() +
     theme(legend.position = "none") +
ggtitle("Varroa mite ploidy
Flow-cytometry") +
   facet_grid(.~stage_mature, space = 'free',drop = F,
              labeller = my_strip_labels) # add labels 

#density plot with rug
p_ploidy_dens_rug = data %>% dplyr::filter(body.part == "Body") %>%
    ggplot(aes(x=Ploidy, color = Sex, fill = Sex)) + 
   geom_density(alpha=.2) +
   geom_rug(aes(x=Ploidy,color = Sex), alpha=0.9, size=1.5,length = unit(0.05, "npc"))+
      ylab("Density") +
      theme_classic2() +
      theme(legend.position = "none") +
      ggtitle("Varroa mite ploidy
Flow-cytometry") +
    facet_grid(.~stage_mature, space = 'free',drop = F,
               labeller = my_strip_labels) # add labels 
```
