---
title: "Supplementary Information: Figures and Analysis for Varroa Destructor Inheritance Study"
author: "Nurit Eliash and Sasha Mikheyev"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    code_folding: hide
    theme: cerulean
    toc: yes
    toc_depth: 5
    toc_float:
      collapsed: yes
      df_print: paged
bibliography: references.bib
editor_options: 
  chunk_output_type: console
  markdown: 
    wrap: 72
---

```{=html}
<style type="text/css">
  body{
  font-size: 12pt;
}
</style>
```
# Overview

This document contains supplementary figures and analyses for our study
on the mode of genetic inheritance in *Varroa destructor*. It should be
read in conjunction with the detailed methodology presented in [Site
Quality Analysis](site_quality_exploration.html).

## Document Structure

1.  Prerequisites and Setup
    -   Project overview
    -   Data sources
    -   Required files and packages
2.  Data Processing
    -   VCF file processing
    -   Site quality filtering methodology
3.  Main Figures
    -   Figure 1: Genotype flow in single family
    -   Figure 2: Cross-family analysis
4.  Supplementary Analyses
    -   Quality metrics
    -   Additional validations

## Data Availability

-   Raw sequencing data: [BioProject
    PRJNA794941](https://www.ncbi.nlm.nih.gov/bioproject/PRJNA794941/)
-   Analysis code: [GitHub
    Repository](https://github.com/nurit-eliash/varroa-pedigree-study/)
-   Filtered datasets: Available upon request

# Prerequisites and Setup

## Project Overview

This analysis is part of a comprehensive study investigating the genetic
inheritance patterns in *Varroa destructor* mites. We analyzed genomic
data from 223 mite samples across 30 families, collected from three
*Apis mellifera* colonies. Each family consists of three generations:

## Data Sources

-   Raw fastq reads: [BioProject
    PRJNA794941](https://www.ncbi.nlm.nih.gov/bioproject/PRJNA794941/)
-   Code repository:
    [GitHub](https://github.com/nurit-eliash/varroa-pedigree-study/tree/main#varroa-pedigree-study)

## Required Files

Before running this analysis, ensure you have:\
1. The VCF file processed with the specified parameters\
2. The filtered sites CSV file generated from site quality exploration\
3. All required R packages installed

1.  **VCF File**: Filtered variant calls meeting these criteria:
    -   Minimum quality score: 15000
    -   Minimum depth: 16
    -   Maximum depth: 40
    -   Maximum missing data: 50%
    -   Minimum allele frequency: 0.2
    -   Chromosomes: NW_019211454.1 through NW_019211460.1

```{r libraries, echo=FALSE, eval=TRUE, warning=FALSE, message=FALSE}
library("tidyverse")
library("dplyr")
library("ggplot2")
library("ggpubr")
library("scales")
library("ggpubr")
library("gridExtra")
library("grid")
library("GGally")
library("vcfR") # for extracting genotype data from a vcf file
library("janitor")
library("splitstackshape")
library("plotly")
library("kableExtra")
library("pastecs") # for stats stat.desc()
library("patchwork")
library("readr")  # for read_delim function
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                     fig.width = 10,
                      fig.asp = 0.4,
                      out.width = "100%")
```

# Data Processing

## VCF File Processing

The initial variant calling file was filtered using VCFtools with
stringent quality control parameters to ensure high-confidence variant
calls. The filtering criteria were:

``` bash
vcftools --vcf snponly_freebayes.vcf \
  --chr NW_019211454.1 \
  --chr NW_019211455.1 \
  --chr NW_019211456.1 \
  --chr NW_019211457.1 \
  --chr NW_019211458.1 \
  --chr NW_019211459.1 \
  --chr NW_019211460.1 \
  --max-alleles 2 \
  --minQ 15000 \
  --minDP 16 \
  --maxDP 40 \
  --max-missing 0.5 \
  --maf 0.2 \
  --recode \
  --recode-INFO-all \
  --out Q15000BIALLDP16HDP40mis.5maf.2Chr7
```

### Filtering Rationale

-   **Chromosomes**: Selected the 7 largest chromosomes
    (NW_019211454.1-460.1)
-   **Alleles**: Limited to biallelic sites to focus on clear binary
    inheritance patterns
-   **Quality**: Minimum Phred score of 15000 ensures high base quality
-   **Depth**: Range of 16-40 balances between coverage requirements and
    potential artifacts
-   **Missing Data**: Maximum 50% missing data per site maintains
    statistical power
-   **Minor Allele Frequency**: Minimum 0.2 reduces impact of sequencing
    errors

## Quality Control Metrics

We implemented two key metrics to assess genotype quality:

1.  **Variant Allele Frequency (VAF)**
    -   Calculated as: VAF = AO/DP
    -   AO: Alternate allele observations
    -   DP: Total read depth
    -   Provides measure of allelic balance
2.  **Quality Allele Frequency (QAF)**
    -   Calculated as: QAF = QA/(QA+QO)
    -   QA: Quality scores for alternate allele
    -   QO: Quality scores for reference allele
    -   Weights allele calls by base quality

### Genotype Classification Thresholds

Based on empirical distribution analysis (see Figures 1-2), we
established the following thresholds:

| Genotype | VAF Range | QAF Range | Interpretation                 |
|----------|-----------|-----------|--------------------------------|
| AA (Ref) | \< 0.03   | \< 0.03   | Strong reference support       |
| AB (Het) | 0.4 - 0.6 | 0.4 - 0.6 | Balanced allele representation |
| BB (Alt) | \> 0.8    | \> 0.8    | Strong alternate support       |

These thresholds were validated using known inheritance patterns in our
pedigree structure.

```{r process-vcf, echo=FALSE, eval=TRUEç}
# Read and process VCF file
vcf <- read.vcfR("/Users/nuriteliash/Documents/GitHub/varroa-pedigree-study/data/vcf_filter/minQ_filter/Q15000BIALLDP16HDP40mis.5maf.2Chr7.recode.vcf", verbose = FALSE)

# Extract genotype data
gt <- extract.gt(vcf, element = "GT") 
gt <- as.data.frame(t(gt)) %>%
  rownames_to_column("ID")

# Clean the ID names
gt$ID <- sub("_[^_]+$", "", gt$ID)

# Create the table object
table <- gt %>%
  t() %>%
  as.data.frame() %>%
  row_to_names(row_number = 1) %>%
  dplyr::select(contains(c("son", "dat", "fnd"))) %>%  # Keep only F0, F1, and F2 adults
  # Remove columns containing specific substrings
  dplyr::select(-matches("sis|479-2|565-2|535_1|535_3")) %>%
  # Convert GT values from numbers to letters
  mutate(across(everything(), ~ case_when(
    . == "0/0" ~ "AA",
    . == "0/1" ~ "AB",
    . == "1/1" ~ "BB",
    TRUE ~ .  # Keep other values unchanged
  )))
```

## Site Quality Filtering

The site quality filtering criteria were developed based on Variant
Allele Frequency (VAF) and Genotype Quality Allele Frequency (QAF)
analysis. For detailed methodology and validation of these thresholds,
see [Site Quality
Analysis](site_quality_exploration.html#sanity-check-to-determine-the-threshold-for-site-filtering).

We applied two additional filtering parameters to account for biases in
alternate allele representation and to ensure high-confidence genotype
calls based on allele frequency and quality.\
*Variant allele frequency (VAF)* was calculated as VAF = AO/DP, where AO
(Alternate Observations) represents the number of reads supporting the
alternate allele, and DP (Depth) is the total read depth at the site.\
Similarly, *Quality Allele Frequency (QAF)* was computed as QAF =
QA/QA+QO, where QA (Quality of Alternate Allele) is the sum of
Phred-scaled base quality scores for reads supporting the alternate
allele, and QO (Quality of Reference Allele) is the sum of Phred-scaled
base quality scores for reads supporting the reference allele.

```{r load-f2-homo-parents}
# Read the f2_homo_parents data (sanity check data) from CSV: F2 females at the filtered positions, on chrom 1, where F1 male and female are homozygous (AA)
# the file was created using the ~/Documents/GitHub/varroa-pedigree-study/R_scripts/After_revision/site_quality_exploration.Rmd
f2_homo_parents <- read.csv("/Users/nuriteliash/Documents/GitHub/varroa-pedigree-study/data/f2_homo_parents.csv")
```

### Supplementary Figure: Distribution of VAF and QAF in F2 Offspring

*Click the code button to view the analysis of VAF and QAF
distributions*

```{r vaf-qaf-distribution, echo=FALSE, eval=TRUE}
# QAF: All genotypes overlaid
QAF_all_gt <- ggplot(f2_homo_parents, aes(x = QAF, fill = GT)) +
  geom_histogram(binwidth = 0.02, alpha = 1, position = "identity") +
  scale_x_continuous(breaks = seq(0, 1, by = 0.1)) +
  theme_classic() +
  labs(title = "A",
       x = "Quality Allele Frequency (QAF)",
       y = "Count") +
  scale_fill_manual(
    values = c("AA" = "#ffbf00", "AB" = "#66b032", "BB" = "#1982c4"),
    name = "Genotype"
  ) +
  theme(
    legend.position = "none",
    axis.text.x = element_text(size = 10, angle = 45, hjust = 1, vjust = 1),
    axis.title.x = element_text(margin = margin(t = 15)),
    plot.margin = margin(10, 10, 30, 10)
  )

# QAF: Facetted by genotype
QAF_facet_gt <- ggplot(f2_homo_parents, aes(x = QAF, fill = GT)) +
  geom_histogram(binwidth = 0.02, alpha = 1, position = "identity") +
  facet_wrap(~ GT, scales = "free_y") +
  scale_x_continuous(breaks = seq(0, 1, by = 0.1)) +
  theme_classic() +
  labs(title = "B",
       x = "Quality Allele Frequency (QAF)",
       y = "Count") +
  scale_fill_manual(
    values = c("AA" = "#ffbf00", "AB" = "#66b032", "BB" = "#1982c4"),
    name = "Genotype"
  ) +
  theme(
    legend.position = "right",
    axis.text.x = element_text(size = 10, angle = 45, hjust = 1, vjust = 1),
    axis.title.x = element_text(margin = margin(t = 15)),
    plot.margin = margin(10, 10, 30, 10),
    panel.spacing = unit(2, "lines")
  )

# Combine QAF plots with adjusted spacing
QAF_combined <- (QAF_all_gt + QAF_facet_gt + plot_layout(widths = c(1, 3)))

# Display the combined QAF plot
QAF_combined

# Add caption below the plot
cat("**Figure 1.** Distribution of Quality Allele Frequency (QAF) values in F2 females from crosses with homozygous F0 and F1 parents.\nA. Overlaid histogram of QAF values for all genotypes.\nB. Same data faceted by genotype (AA, AB, BB), showing the distribution and spread within each genotype class.\nVertical axis represents count of sites; horizontal axis is QAF. Bin width = 0.02.\nGenotypes are color-coded: AA (yellow), AB (green), BB (blue).")

# VAF: All genotypes overlaid
VAF_all_gt <- ggplot(f2_homo_parents, aes(x = VAF, fill = GT)) +
  geom_histogram(binwidth = 0.02, alpha = 1, position = "identity") +
  scale_x_continuous(breaks = seq(0, 1, by = 0.1)) +
  theme_classic() +
  labs(title = "A",
       x = "Variant Allele Frequency (VAF)",
       y = "Count") +
  scale_fill_manual(
    values = c("AA" = "#ffbf00", "AB" = "#66b032", "BB" = "#1982c4"),
    name = "Genotype"
  ) +
  theme(
    legend.position = "none",
    axis.text.x = element_text(size = 10, angle = 45, hjust = 1, vjust = 1),
    axis.title.x = element_text(margin = margin(t = 15)),
    plot.margin = margin(10, 10, 30, 10)
  )

# VAF: Facetted by genotype
VAF_facet_gt <- ggplot(f2_homo_parents, aes(x = VAF, fill = GT)) +
  geom_histogram(binwidth = 0.02, alpha = 1, position = "identity") +
  facet_wrap(~ GT, scales = "free_y") +
  scale_x_continuous(breaks = seq(0, 1, by = 0.1)) +
  theme_classic() +
  labs(title = "B",
       x = "Variant Allele Frequency (VAF)",
       y = "Count") +
  scale_fill_manual(
    values = c("AA" = "#ffbf00", "AB" = "#66b032", "BB" = "#1982c4"),
    name = "Genotype"
  ) +
  theme(
    legend.position = "right",
    axis.text.x = element_text(size = 10, angle = 45, hjust = 1, vjust = 1),
    axis.title.x = element_text(margin = margin(t = 15)),
    plot.margin = margin(10, 10, 30, 10),
    panel.spacing = unit(2, "lines")
  )

# Combine VAF plots with adjusted spacing
VAF_combined <- (VAF_all_gt + VAF_facet_gt + plot_layout(widths = c(1, 3)))

# Display the combined VAF plot
VAF_combined

# Add caption below the plot
cat("**Figure 2.** Distribution of Variant Allele Frequency (VAF) values in F2 females from crosses with homozygous F0 and F1 parents.\nA. Overlaid histogram of VAF values across all genotypes.\nB. Faceted distribution of VAF by genotype, showing characteristic frequency patterns for AA, AB, and BB.\nVertical axis represents count of sites; horizontal axis is VAF. Bin width = 0.02.\nGenotypes are color-coded: AA (yellow), AB (green), BB (blue).")
```

Based on the distribution of sites' VAF and QAF values, the following
filtering criteria were applied:\
- Homozygous Reference (AA): VAF \< 0.03 and QAF \< 0.03, sites exhibit
low alternate allele support\
- Heterozygous (AB): 0.4 \< VAF \< 0.6 and 0.4 \< QAF \< 0.6, sites
maintain a balanced allele distribution\
- Homozygous Alternate (BB): VAF \> 0.8 and QAF \> 0.8, sites
demonstrate high alternate allele support

This filtering approach enhances the accuracy of genotype classification
while reducing the inclusion of low-confidence variant calls, by
minimizing genotyping errors due to sequencing artifacts or
depth-related biases.

# Main Figures

## Figure 1A: Crossing F1 mother AB with F1 father AA, one family

*Note: The original analysis can be found in:* -
`~/Documents/GitHub/varroa-pedigree-study/R_scripts/After_revision/site_quality_exploration.Rmd`

```{r load-selected-family}
# Load the fam2_selected data, for one family (478) on the largest chromosome, NW_019211454
fam2_selected <- read.csv("/Users/nuriteliash/Documents/GitHub/varroa-pedigree-study/data/fam2_selected.csv")
```

```{r plot-one-family}
# Create a numeric x-axis for spacing (convert 'member' to factor and assign spacing)
fam_id <- 478

fam2_selected %>%
  ggplot(aes(x = member_num, y = pos / 1e6, color = GT)) +  # Convert pos to Mb
  geom_segment(aes(x = member_num - 0.3, xend = member_num + 0.3, 
                   y = pos / 1e6, yend = pos / 1e6), size = 1.5) +  # Apply same conversion
  scale_x_continuous(breaks = unique(fam2_selected$member_num), labels = unique(fam2_selected$member)) +  # Restore original labels
  scale_color_manual(
    values = c("AA" = "#ffbf00", "AB" = "#66b032", "BB" = "#1982c4"),  # Updated colors for renamed genotypes
    name = "Genotype",
    breaks = c("AA", "AB", "BB")  # Ensures BB appears in the legend even if absent
  ) +
  theme_classic() +
  labs(
    title = paste("Genotype Flow Across Family Members\nFamily ID:", fam_id),
    x = "Family Member",
    y = "Genomic Position (Mb)"
  ) +
  theme(
    axis.text.x = element_text(size = 10, angle = 45, hjust = 1, vjust = 1),  # Adjusted text size and angle
    axis.title.x = element_text(margin = margin(t = 15)),  # Add margin above x-axis title
    panel.grid.major = element_blank(), 
    legend.position = "right",  # Move legend to the right
    legend.box.margin = margin(0, 0, 0, 10),  # Add margin to separate legend from plot
    plot.margin = margin(10, 20, 30, 10)  # Increased bottom margin to accommodate rotated labels
  )
```

## Figure 2B: Crossing F1 mother AB with F1 father AA, all families

```{r filtered-sites, echo=FALSE, eval=TRUE}
# Set the families (include only families with at least one adult F2)
family <- grep("grndat|grnson", gt$ID, value=TRUE) %>%
  str_extract("[^_]+") %>%
  unique()

# Define a list to store all data frames per family
obs <- list()

for (fam in family) {
    obs[[fam]] <- table %>%
      dplyr::select(starts_with(fam)) %>%
    # Add site and chromosome information
      rownames_to_column("site") %>%
    mutate(chrom = str_split(site, pattern = "_", simplify = TRUE)[, 2]) %>%
    # Filter for the genotype patterns we want
      dplyr::filter_at(vars(matches("_fnd")), all_vars(. == "AA")) %>%  # F0 Female must be AA
      dplyr::filter_at(vars(matches("_son")), all_vars(. == "AA")) %>%  # Sons must be AA
      dplyr::filter_at(vars(matches("_dat")), all_vars(. == "AB")) %>%  # Daughters must be AB
    # Select only the grandchildren columns and keep site/chrom info
    dplyr::select(site, chrom, contains("grn")) %>%
    # Reshape the data
      tidyr::pivot_longer(cols = -c(site, chrom)) %>%
    dplyr::rename(Sample = name, gt = value) %>%
    # Count genotypes
    dplyr::count(Sample, gt, chrom, .drop = FALSE) %>%
      dplyr::filter(gt %in% c("AA", "BB", "AB")) %>%
      mutate(n = as.numeric(n)) %>%
    group_by(Sample) %>%
      mutate(total = as.numeric(sum(n))) %>%
      mutate(sex = case_when(
      grepl("son", Sample) ~ "Male",
      grepl("dat", Sample) ~ "Female"
      )) %>%
    # Add family column
      mutate(family = fam)
}

# Bind all families together into a final data frame containing all observed counts
observed <- do.call("rbind", obs) %>% 
  mutate(Sample = as.character(Sample))

# Create samples dataframe
samples <- data.frame(
  Sample = rep(unique(observed$Sample), each = 3), 
  gt = rep(c("AA", "AB", "BB"))
) %>%   
  mutate(family = str_extract(Sample, "^[0-9]+"))

# Join and create final dataset
samples_obs_all <- left_join(samples, observed, by = c("family", "Sample", "gt")) %>% 
  mutate(sex = case_when(
    grepl("son", Sample) ~ "Male",
    grepl("dat", Sample) ~ "Female"
  )) %>%
  group_by(Sample, chrom) %>%
    mutate(total = as.numeric(sum(n))) %>%
    dplyr::mutate(prop = n/total) %>%
    na.omit()
```

### Analysis of filtered sites (site quality QAF and VAF)

```{r load-filtered-sites}
# Load the pre-filtered sites per family and sample
# the file was created using the ~/Documents/GitHub/varroa-pedigree-study/R_scripts/After_revision/site_quality_exploration.Rmd

filtered_sites <- read_csv("/Users/nuriteliash/Documents/GitHub/varroa-pedigree-study/data/filtered_sites.csv") %>%
    mutate(family = as.character(family),
           Sample = as.character(Sample),
           # Extract Chromosome from the site column
           chrom = str_split(site, pattern = "_", simplify = TRUE)[, 2]) %>% 
    rename(gt = GT)
```

```{r filtered-sites-analysis, echo=FALSE, eval=TRUE}
# Set the families (include only families with at least one adult F2)
family <- grep("grndat|grnson", gt$ID, value=TRUE) %>%
  str_extract("[^_]+") %>%
    unique()

# Define a list to store all data frames per family
obs <- list()

for (fam in family) {
obs[[fam]] <- table %>%
  dplyr::select(starts_with(fam)) %>%
    # Add site and chromosome information
    rownames_to_column("site") %>%
    mutate(chrom = str_split(site, pattern = "_", simplify = TRUE)[, 2]) %>%
    # Then filter for the genotype patterns we want
    dplyr::filter_at(vars(matches("_fnd")), all_vars(. == "AA")) %>%  # F0 Female must be AA
    dplyr::filter_at(vars(matches("_son")), all_vars(. == "AA")) %>%  # Sons must be AA
    dplyr::filter_at(vars(matches("_dat")), all_vars(. == "AB")) %>%  # Daughters must be AB
    # Select only the grandchildren columns and keep site/chrom info
    dplyr::select(site, chrom, contains("grn")) %>%
    # Reshape the data
    tidyr::pivot_longer(cols = -c(site, chrom)) %>%
    dplyr::rename(Sample = name, gt = value) %>%
    # filter for sites 
    left_join(filtered_sites, by = c("site", "chrom", "Sample", "gt")) %>%
    na.omit() %>% # Count genotypes
    dplyr::count(Sample, gt, chrom, .drop = FALSE) %>%
  dplyr::filter(gt %in% c("AA", "BB", "AB")) %>%
  mutate(n = as.numeric(n)) %>%
    group_by(Sample) %>%
  mutate(total = as.numeric(sum(n))) %>%
  mutate(sex = case_when(
      grepl("son", Sample) ~ "Male",
      grepl("dat", Sample) ~ "Female"
    )) %>%
    # Add family column
    mutate(family = fam)
}

# Bind all families together into a final data frame containing all observed counts
observed <- do.call("rbind", obs) %>% 
  mutate(Sample = as.character(Sample))

# Create samples dataframe
samples <- data.frame(
  Sample = rep(unique(observed$Sample), each = 3), 
  gt = rep(c("AA", "AB", "BB"))
) %>%   
  mutate(family = str_extract(Sample, "^[0-9]+"))

# Join and create final dataset
samples_obs_filtered <- left_join(samples, observed, by = c("family", "Sample", "gt")) %>% 
  mutate(sex = case_when(
    grepl("son", Sample) ~ "Male",
    grepl("dat", Sample) ~ "Female"
  )) %>%
  group_by(Sample, chrom) %>%
    mutate(total = as.numeric(sum(n))) %>%
    dplyr::mutate(prop = n/total) %>%
  na.omit()
```

### Summary of sites filtrered in each sample

```{r summary-filtered-sites, echo=FALSE, eval=TRUE}
# Create a table showing sites filtered per sample
sites_per_sample <- table %>%
  rownames_to_column("site") %>%
  tidyr::pivot_longer(cols = -site, names_to = "Sample", values_to = "gt") %>%
  # Filter out NA values before counting
  filter(!is.na(gt)) %>%
  group_by(Sample) %>%
  summarise(
    total_sites = n(),
    .groups = 'drop'
  )

filtered_sites_per_sample <- filtered_sites %>%
  group_by(Sample) %>%
  summarise(
    filtered_sites = n(),
    .groups = 'drop'
  )

# Join and calculate differences
sites_comparison <- sites_per_sample %>%
  left_join(filtered_sites_per_sample, by = "Sample") %>%
  mutate(
    filtered_sites = replace_na(filtered_sites, 0),
    sites_removed = total_sites - filtered_sites,
    percent_removed = round((sites_removed / total_sites) * 100, 2)
  ) %>%
  arrange(desc(sites_removed))

# Display the table
sites_comparison %>%
  select(Sample, total_sites, filtered_sites, sites_removed, percent_removed) %>%
  rename(
    "Sample" = Sample,
    "Total Sites" = total_sites,
    "Sites Kept" = filtered_sites,
    "Sites Removed" = sites_removed,
    "% Removed" = percent_removed
  ) %>%
  knitr::kable(
    digits = 2,
    caption = "Number of sites filtered per sample",
    format = "html"
  ) %>%
  kableExtra::kable_styling(
    bootstrap_options = "striped",
    full_width = TRUE,
    position = "left",
    fixed_thead = TRUE
  ) %>%
  kableExtra::scroll_box(
    width = "100%", 
    height = "300px"
  )

# Add a note about scrolling below the table
cat("\n*Note: Scroll to see all rows in the table.*")
```

### Filtered for QAF and VAF, minimum sites per sample = 40

*Click the code button to view the analysis with minimum 40 sites per
sample*

```{r filtered-data-sites-min40, echo=TRUE, eval=TRUE, message=FALSE}
# filtered samples, at least 40 sites per sample
site_threshold = as.numeric(40)

## filtered sites

# Create summary dataframe for labels, summing across all chromosomes (for pulled data only)
summary_df_filtered <- samples_obs_filtered %>%
  group_by(sex) %>%  # Remove chromosome grouping
  summarise(
    num_samples = n_distinct(Sample),  # Count unique samples
    total_sites = sum(total, na.rm = TRUE),  # Sum all sites
    .groups = 'drop'
  )

# Create per-chromosome summary for faceted plot
summary_df_filtered_by_chrom <- samples_obs_filtered %>%
  group_by(chrom, sex) %>%
  summarise(
    num_samples = n_distinct(Sample),
    total_sites = sum(total, na.rm = TRUE),
    .groups = 'drop'
  )

# Plot 1: Pulled data with totals across all chromosomes
ggplot(samples_obs_filtered %>% filter(total >= site_threshold), 
       aes(fill = gt, y = prop, x = factor(sex, levels = c("Female", "Male")))) +
  geom_bar(position = "fill", stat = "identity", width = 0.5) +
  ylab("Genotype Proportion") +
  labs(title = "Offspring Genotype, Crossing (AA) Male x (AB) Female
 Filtered sites, sites per sample => 40", 
       fill = "Genotype") +
  theme_classic() +
  theme(
    axis.title.x = element_blank(),
    axis.text.x = element_blank(),
    axis.ticks.x = element_blank(),
    axis.text.y = element_text(size = 12),
    strip.text.x = element_text(size = 12),
    panel.spacing = unit(1, "lines"),
    plot.margin = margin(t = 10, r = 10, b = 60, l = 10)
  ) +
  scale_fill_manual(values = c("#ffbf00", "#66b032", "#1982c4")) +
  # Add sample count labels (only for pulled data)
  geom_text(data = summary_df_filtered,
            aes(x = sex, 
                y = -0.03,
                label = paste0(num_samples, " ", sex, "s")),
            inherit.aes = FALSE,
            size = 2.8,
            vjust = 1) +
  # Add total sites label with comma formatting (only for pulled data)
  geom_text(data = summary_df_filtered,
            aes(x = sex, 
                y = -0.18,
                label = paste0(format(total_sites, big.mark = ","), " sites")),
            inherit.aes = FALSE,
            size = 2.8,
            vjust = 0.5)

# Plot 2: Faceted plot by chromosome (with per-chromosome statistics)
ggplot(samples_obs_filtered %>% filter(total >= site_threshold), 
       aes(fill = gt, y = prop, x = factor(sex, levels = c("Female", "Male")))) +
  geom_bar(position = "fill", stat = "identity", width = 0.5) +
  facet_wrap(~chrom, nrow = 1) +
  ylab("Genotype Proportion") +
  labs(title = "Offspring Genotype, Crossing (AA) Male x (AB) Female
 Filtered sites, sites per sample => 40", 
       fill = "Genotype") +
  theme_classic() +
  theme(
    axis.title.x = element_blank(),
    axis.text.x = element_blank(),
    axis.ticks.x = element_blank(),
    axis.text.y = element_text(size = 12),
    strip.text.x = element_text(size = 12),
    panel.spacing = unit(1.5, "lines"),  # Increased panel spacing
    plot.margin = margin(t = 10, r = 10, b = 60, l = 10),
    legend.position = "none"  # Remove legend
  ) +
  scale_fill_manual(values = c("#ffbf00", "#66b032", "#1982c4")) +
  # Add sample count labels per chromosome
  geom_text(data = summary_df_filtered_by_chrom,
            aes(x = sex, 
                y = -0.03,
                label = paste0(num_samples, " ", sex, "s")),
            inherit.aes = FALSE,
            size = 2.8,
            vjust = 1) +
  # Add sites count labels per chromosome
  geom_text(data = summary_df_filtered_by_chrom,
            aes(x = sex, 
                y = -0.18,
                label = paste0(format(total_sites, big.mark = ","), " sites")),
            inherit.aes = FALSE,
            size = 2.8,
            vjust = 0.5)

# Plot 3: Box plot remains unchanged
ggplot(samples_obs_filtered %>% 
       filter(total >= site_threshold) %>%
       filter(gt == "AA"), 
       aes(x = factor(sex, levels = c("Female", "Male")), y = prop, color = sex)) +
  geom_boxplot(fill = "grey80", outlier.shape = NA, color = "black") +
  geom_point(position = position_jitter(width = 0.2), size = 2, alpha = 0.7) +
  facet_wrap(~chrom, nrow = 2) +
  ylab("Proportion of AA Genotype") +
  xlab("Sex") +
  labs(title = "Distribution of AA Genotype Proportion per Chromosome
Filtered sites, sites per sample => 40") +
  theme_classic() +
  theme(
    strip.background = element_rect(fill = "white", color = "black"),
    axis.text = element_text(size = 12),
    strip.text = element_text(size = 12),
    panel.spacing = unit(1, "lines"),
    plot.margin = margin(t = 10, r = 10, b = 10, l = 10)
  ) +
  scale_color_manual(values = c("Female" = "red", "Male" = "blue")) +
  coord_cartesian(ylim = c(0, 1))
```

# Supplementary Figures for Manuscript

## Variant Quality Distribution (Figure S 6)

```{r variant-quality-dist, echo=FALSE, eval=TRUE, warning=FALSE, message=FALSE}
# Read the variant quality data
var_qual <- read_delim("/Users/nuriteliash/Documents/GitHub/varroa-pedigree-study/data/vcf_filter/Q40BIALLDP16HDP40mis.5Chr7/Q40BIALLDP16HDP40mis.5Chr7.lqual", 
                       delim = "\t",
                       col_names = c("chr", "pos", "qual"), 
                       skip = 1)

# Print summary statistics
cat("Summary of variant quality scores:\n")
print(summary(var_qual$qual))

# Create main plot
ggplot(var_qual, aes(x = qual)) +
  geom_density(fill = "dodgerblue1", colour = "black", alpha = 0.3) + 
  theme_classic() +
  ggtitle("Variant Quality Distribution") +
  scale_x_continuous(n.breaks = 12, limits = c(0, 50000)) +  # Adjusted x-axis range
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    plot.title = element_text(size = 12),
    axis.title = element_text(size = 10)
  ) +
  xlab("Variant Quality") + 
  ylab("Density") +
  geom_vline(xintercept = 15000, color = "red", linetype = "dashed")

# Calculate percentage of variants above threshold
pct_above_threshold <- mean(var_qual$qual >= 15000) * 100

# Add caption below the plot
cat(sprintf("\n**Figure S6.** Distribution of variant quality scores across sites that passed initial filtering (depth ≥16, ≤40; maximum 50%% missing data). The red dashed line indicates the quality threshold (15000) used for final filtering. %.1f%% of variants have quality scores above this threshold. Higher quality scores indicate greater confidence in variant calls.", pct_above_threshold))
```

## Karyotyping of males and females (Figure S 3A)

```{r karyotyping-plot, echo=TRUE, eval=FALSE}
# Code for karyotyping plot will be added here
# This will generate Figure S 3A showing karyotype analysis of males and females
```

## Flow cytometry of males and females (Figure S 3B)

```{r flow-cytometry-plot, echo=TRUE, eval=FALSE}
# Code for flow cytometry plot will be added here
# This will generate Figure S 3B showing flow cytometry analysis of males and females
```

## RNAseq of males and females (Figure 2)

```{r rnaseq-plot, echo=TRUE, eval=FALSE}
# Code for RNAseq analysis plot will be added here
# This will generate Figure 2 showing RNAseq analysis of males and females
```

------------------------------------------------------------------------

# Archived Code for Future Reference

*The following code sections are preserved for future analysis but are
not visible in the HTML output.*

```{r archived-code-min10, eval=FALSE, include=FALSE}
### Filtered for QAF and VAF, minimum sites per sample = 10
site_threshold = as.numeric(10)

## all sites
# First, create the summary dataframe for labels
summary_df_all <- samples_obs_all %>%
  group_by(chrom, sex) %>%
  summarise(
    num_samples = length(unique(Sample)),
    total_sites = sum(total, na.rm = TRUE),
    .groups = 'drop'
  )

# plot pulled data
ggplot(samples_obs_all %>% filter(total >= site_threshold), 
       aes(fill = gt, y = prop, x = factor(sex, levels = c("Female", "Male")))) +
  geom_bar(position = "fill", stat = "identity", width = 0.5) +
  ylab("Genotype Proportion") +
  labs(title = "Offspring Genotype, Crossing (AA) Male x (AB) Female
All sites, sites per sample => 10", 
       fill = "Genotype") +
  theme_classic() +
  theme(
    axis.title.x = element_blank(),
    axis.text.x = element_blank(),
    axis.ticks.x = element_blank(),
    axis.text.y = element_text(size = 12),
    strip.text.x = element_text(size = 12),
    panel.spacing = unit(1, "lines"),
    plot.margin = margin(t = 10, r = 10, b = 60, l = 10)
  ) +
  scale_fill_manual(values = c("#ffbf00", "#66b032", "#1982c4")) +
  geom_text(data = summary_df_all,
            aes(x = sex, 
                y = -0.03,
                label = paste0(num_samples, " ", sex, "s")),
            inherit.aes = FALSE,
            size = 2.8,
            vjust = 1) +
  geom_text(data = summary_df_all,
            aes(x = sex, 
                y = -0.18,
                label = paste0(format(total_sites, big.mark = ","), " sites")),
            inherit.aes = FALSE,
            size = 2.8,
            vjust = 0.5)

# [Rest of the archived code for min10 analysis...]
```

```{r archived-code-f1-male-ab, eval=FALSE, include=FALSE}
## F1 male (AB) x Female (AB)
# [All the F1 male (AB) x Female (AB) analysis code...]
```

```{r archived-code-f1-male-ab-female-aa, eval=FALSE, include=FALSE}
## F1 male (AB) x Female (AA)
# [All the F1 male (AB) x Female (AA) analysis code...]
```

```{r archived-code-heterozygosity, eval=FALSE, include=FALSE}
## Heterozygosity Analysis
# [All the heterozygosity analysis code...]
```

```{r archived-code-ploidy, eval=FALSE, include=FALSE}
## Ploidy Analysis
# [All the ploidy analysis code...]
```
