---
title: "Varroa genotype distribution"
author: "Nurit Eliash"
date: "May 2023"
output:
  html_document:
    code_folding: hide
    theme: cerulean
    toc: yes
    toc_depth: 5
    toc_float:
      collapsed: yes
      df_print: paged
editor_options: 
  chunk_output_type: console
---

```{=html}
<style type="text/css">
  body{
  font-size: 12pt;
}
</style>
```

## load libraries
```{r libraries, echo=TRUE, message=FALSE, warning=FALSE}
library("tidyverse")
library("plyr")
library("dplyr")
library("ggplot2")
library("scales")
library("ggpubr")
library("gridExtra")
library("grid")
library("GGally")
library("vcfR") # for extracting genotype data from a vcf file
library("data.table")
library("stringr")
library("janitor")
library("plotly") # for the 3d plots
library("ggfortify")
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                     fig.width = 10,
                      fig.asp = 0.6,
                      out.width = "100%")
```

## Load Variant Call Format (VCF) file.
Extract genotypes for each site and individual. The metadata for all samples can be found in [here](https://github.com/nurit-eliash/varroa-linkage-map/blob/main/data/meta_data_223.xlsx).

```{r}
vcf <- read.vcfR("/Users/nuriteliash/Documents/GitHub/varroa-linkage-map/data/vcf_filter/Q40BIALLDP16HDP40mis.5Chr7/Q40BIALLDP16HDP40mis.5Chr7.recode.vcf", verbose = FALSE )
vcf
# extract the genotype for each site in each individual
gt <- extract.gt(vcf, element = "GT") 
gt <- as.data.frame(t(gt)) %>%
  rownames_to_column("ID")
#clean the ID names
gt$ID <- sub("_[^_]+$", "", gt$ID)

table <-  gt %>% 
  t() %>%
  as.data.frame() %>%
  row_to_names(row_number = 1) # %>%
 # dplyr::select(-c("48_49e_grn", "48_49f_grn", "48_49g_grn")) # remove samples: 48_49e_grn, 48_49f_grn, 48_49g_grn. they were born in a second cycle, and in the first cycle an adult male was born in F2. so i cannot tell for sure who is the father 
```

read the meta data table and set the family
```{r}
meta <- read_csv("/Users/nuriteliash/Documents/GitHub/varroa-linkage-map/data/meta_more.csv")

# set the families 
family = grep("fnd",gt$ID, value=TRUE) %>%
  str_extract("[^_]+")  %>%
  unique()
```

## Control: Homozygotic sites (crosses 1-4)
### (1) F1 male (0/0) x female (0/0)

we expect all F2 offspring to be homozygotic (0/0) like their parents (F1)
```{r}
# Prepare table with observed and expected counts, per family:
# define a list to put all the data frames in
obs <- list()

for (fam in family) {
 
obs[[fam]] <- table %>%
  dplyr::select(starts_with(fam)) %>%
        dplyr::filter_at(vars(matches("_fnd")), all_vars(. == "0/0")) %>% # force F0 female to be homo, like her son
dplyr::filter_at(vars(matches("_son")), all_vars(. == "0/0")) %>%
  dplyr::filter_at(vars(matches("_dat")), all_vars(. == "0/0")) %>% 
  dplyr::select(contains("grn")) %>%
  tidyr::pivot_longer(everything())  %>% 
  dplyr::rename(sample = name, gt = value) %>%
  dplyr::count(sample, gt, .drop = FALSE) %>% 
  dplyr::filter(gt %in% c("0/0", "1/1", "0/1")) %>%
  mutate(n = as.numeric(n)) %>%
  group_by(sample) %>%
#  mutate(total = as.numeric(sum(n))) %>%
  dplyr::rename(obs = n)
}

# bind all families together, to a final data frame containing all observed counts
observed <- do.call("rbind", obs) %>% mutate(sample = as.character(sample))
# replace "NA" with zero value when gt doesn't exists
samples <- data.frame(sample = rep(unique(observed$sample), each =3), gt = rep(c("0/0", "0/1", "1/1"))) # add all possible genotypes

data <- left_join(samples, observed, by=c("sample","gt")) %>%  
    replace(is.na(.),0) %>% # replace "NA" with zero value when gt doesn't exists 
    group_by(sample) %>%
    mutate(total = as.numeric(sum(obs))) %>%
    mutate(freq=obs/total) %>% 
    pivot_wider(names_from="gt", values_from=freq, id_cols=c(sample,total)) %>%
    left_join(meta, by="sample") %>%
    rename(homo_ref="0/0") %>%
      rename(hetero="0/1") %>%
    rename(homo_alt="1/1")

head(data)

data %>% 
  plot_ly(x= ~homo_ref, y =~hetero, z = ~homo_alt,
          color = ~sex, 
          colors = c("Egg"="#FFD301","Nymph"="#9BA4B5","female"="#EB455F","male"="#1982c4"),
          symbol= ~adult_sisters,
          symbols= c("with adult sister"="circle", "no adult sister"= "circle-open"),
          text=~sample,
          hoverinfo= "text") %>%
  add_markers() %>%
  layout(title = 'F2 offspring of F1 male (0/0) x female (0/0)',
        scene = list(
        xaxis = list(title = '0/0', nticks = 3, range = c(0,1)),
        yaxis = list(title = '0/1', nticks = 3, range = c(0,1)),
        zaxis = list(title = '1/1', nticks = 3, range = c(0,1))))

# 2d
df <- left_join(samples, observed, by=c("sample","gt")) %>%  
    replace(is.na(.),0) %>% # replace "NA" with zero value when gt doesn't exists 
    group_by(sample) %>%
    mutate(total = as.numeric(sum(obs))) %>%
    mutate(freq=round(obs/total, digit=2)) %>%
  left_join(meta, by="sample")

p = ggplot(df, aes(x = sex, y = freq, colour = gt, shape = adult_sisters,
              text = paste("sample:", sample,
                          ", n:", total))) + 
  geom_boxplot(outlier.shape = NA, coef=0 ) +
  geom_jitter(width=0.1, size=2) +
  scale_color_manual(values=c("#ffbf00", "#66b032","#1982c4")) +
  theme_classic() +
scale_shape_manual(values=c(1,19)) +
  ggtitle('F2 offspring of F1 male (0/0) x female (0/0)') +
    facet_wrap( ~gt)  +  guides(color = "none") 

ggplotly(p) %>% 
  layout(legend = list(orientation = "h", x = 0, y = -0.1)) 
```

```{r eval=FALSE, include=FALSE}
# paying with some plots
 # the old one: with median

# compute median by group
median <- df %>% group_by(gt,sex) %>% summarise(med_val=median(freq))
 
p <- df %>% 
  filter(total>10) %>%
  ggplot(aes(y =freq, x = sex, color = gt, 
              shape = adult_sisters,
              text = paste("sample:", sample,
                          ", n:", total))) +
  geom_jitter(width=0.1, size=2) +
  scale_shape_manual(values=c(1,19)) +
scale_color_manual(values=c("#ffbf00", "#66b032","#1982c4")) +
    theme_classic()  +
  ggtitle('F2 offspring of F1 male (0/0) x female (0/1)') +
  guides(color = "none") +
facet_wrap( ~gt) + 
  geom_hline(data= median, aes(yintercept = med_val))

ggplotly(p) %>% 
  layout(legend = list(orientation = "h", x = 0, y = -0.1))
####
## example
set.seed(8)
y <- rnorm(200)
group <- sample(LETTERS[1:3], size = 200,
                replace = TRUE)
d <- data.frame(y, group)
# Box plot by group with jitter
ggplot(d, aes(x = group, y = y, colour = group)) + 
  geom_boxplot(outlier.shape = NA) +
  geom_jitter()

# no my data





## when i add a text lable, it removes the boxplots...
p <- df %>% 
  filter(total>10) %>%
  ggplot(aes(y =freq, x = sex, color = gt, 
              shape = adult_sisters,
              text = paste("sample:", sample,
                          ", n:", total))) +
  geom_jitter(width=0.1, size=2) +
  scale_shape_manual(values=c(1,19)) +
  scale_color_manual(values=c("#ffbf00", "#66b032","#1982c4")) +
  theme_classic() +
  ggtitle('F2 offspring of F1 male (0/0) x female (0/0)') +
  guides(color = "none") +
    geom_hline(data= median, aes(yintercept = med_val)) +
facet_wrap( ~gt) 

ggplotly(p) %>% 
  layout(legend = list(orientation = "h", x = 0, y = -0.1)) 

# לא הצלחתי להפריד את הקו של הממוצע כך שיהיה קו נפרד לכל סקס. 
# צריך לנסות להפריד, 
# או , לעעשות פעמיים פאסט?
 df %>% 
  filter(total>10) %>%
  ggplot(aes(y =freq, x = sex, color = gt, 
              shape = adult_sisters,
              text = paste("sample:", sample,
                          ", n:", total))) +
  geom_jitter(width=0.1, size=2) +
  scale_shape_manual(values=c(1,19)) +
  scale_color_manual(values=c("#ffbf00", "#66b032","#1982c4")) +
  theme_classic() +
  ggtitle('F2 offspring of F1 male (0/0) x female (0/0)') +
  guides(color = "none") +
facet_wrap( ~gt) 



### 

#this gives too much scattered data... hard to compare sexes
 df %>% 
  filter(total>10) %>%
  ggplot(aes(y =freq, x = sex, color = gt, 
              shape = adult_sisters,
              text = paste("sample:", sample,
                          ", n:", total))) +
  geom_jitter(width=0.1, size=2) +
  scale_shape_manual(values=c(1,19)) +
  scale_color_manual(values=c("#ffbf00", "#66b032","#1982c4")) +
  theme_classic() +
  ggtitle('F2 offspring of F1 male (0/0) x female (0/0)') +
  guides(color = "none") +
    geom_hline(data= median, aes(yintercept = med_val)) +
facet_wrap( ~gt+sex) 
 
##!!!!! *********************
## TRY##
# this gives a median for all sexes, per gt (too litlle lines..)
df %>% 
  filter(total>10) %>%
  ggplot(aes(y =freq, x = sex, color = gt, 
              shape = adult_sisters,
              text = paste("sample:", sample,
                          ", n:", total))) +
  geom_jitter(width=0.1, size=2) +
  scale_shape_manual(values=c(1,19)) +
  scale_color_manual(values=c("#ffbf00", "#66b032","#1982c4")) +
  theme_classic() +
  ggtitle('F2 offspring of F1 male (0/0) x female (0/0)') +
  guides(color = "none") +
facet_wrap( ~gt) +
    geom_hline(data= median, aes(yintercept = med_val, group = sex))

# this option gives a line for each dot... not median (too many lines)
df %>% 
  filter(total>10) %>%
  ggplot(aes(y =freq, x = sex, color = gt, 
              shape = adult_sisters,
              text = paste("sample:", sample,
                          ", n:", total))) +
  geom_jitter(width=0.1, size=2) +
  scale_shape_manual(values=c(1,19)) +
  scale_color_manual(values=c("#ffbf00", "#66b032","#1982c4")) +
  theme_classic() +
  ggtitle('F2 offspring of F1 male (0/0) x female (0/0)') +
  guides(color = "none") +
 #stat_summary(fun = "median", fun.min = "median", fun.max= "median", size= 0.3, geom = "crossbar") +
  stat_summary(fun = "median", geom = "crossbar", group=df$sex)+
facet_wrap( ~gt) 




# this option gives
df %>% 
  filter(total>10) %>%
  ggplot(aes(y =freq, x = sex, color = gt, 
              shape = adult_sisters,
              text = paste("sample:", sample,
                          ", n:", total))) +
  geom_jitter(width=0.1, size=2) +
  scale_shape_manual(values=c(1,19)) +
  scale_color_manual(values=c("#ffbf00", "#66b032","#1982c4")) +
  theme_classic() +
  ggtitle('F2 offspring of F1 male (0/0) x female (0/0)') +
  guides(color = "none") +
 geom_errorbar(data=median, aes(ymin = med_val, ymax = med_val)) +
facet_wrap( ~gt) 


```


### (2) F1 male (1/1) x female (1/1)

we expect all F2 offspring to be homozygotic (1/1) like their parents (F1)
```{r}
# Prepare table with observed and expected counts, per family:
# define a list to put all the data frames in
obs <- list()

for (fam in family) {
 
obs[[fam]] <- table %>%
  dplyr::select(starts_with(fam)) %>%
        dplyr::filter_at(vars(matches("_fnd")), all_vars(. == "1/1")) %>% # force F0 female to be homo, like her son
dplyr::filter_at(vars(matches("_son")), all_vars(. == "1/1")) %>%
  dplyr::filter_at(vars(matches("_dat")), all_vars(. == "1/1")) %>% 
  dplyr::select(contains("grn")) %>%
  tidyr::pivot_longer(everything())  %>% 
  dplyr::rename(sample = name, gt = value) %>%
  dplyr::count(sample, gt, .drop = FALSE) %>% 
  dplyr::filter(gt %in% c("0/0", "1/1", "0/1")) %>%
  mutate(n = as.numeric(n)) %>%
  group_by(sample) %>%
#  mutate(total = as.numeric(sum(n))) %>%
  dplyr::rename(obs = n)
}

# bind all families together, to a final data frame containing all observed counts
observed <- do.call("rbind", obs) %>% mutate(sample = as.character(sample))
# replace "NA" with zero value when gt doesn't exists
samples <- data.frame(sample = rep(unique(observed$sample), each =3), gt = rep(c("0/0", "0/1", "1/1"))) # add all possible genotypes

data <- left_join(samples, observed, by=c("sample","gt")) %>%  
    replace(is.na(.),0) %>% # replace "NA" with zero value when gt doesn't exists 
    group_by(sample) %>%
    mutate(total = as.numeric(sum(obs))) %>%
    mutate(freq=obs/total) %>% 
  mutate(arcsin=asin(sqrt(freq))) %>%
    pivot_wider(names_from="gt", values_from=freq, id_cols=c(sample,total)) %>%
    left_join(meta, by="sample") %>%
    rename(homo_ref="0/0") %>%
      rename(hetero="0/1") %>%
    rename(homo_alt="1/1")

head(data)

data %>% #dplyr::filter(sex == "male") %>%
#  filter(!sample =="502_503d_grn") %>%
#    filter(!sample =="43_44c_grn") %>%
  plot_ly(x= ~homo_ref, y =~hetero, z = ~homo_alt,
          color = ~sex, 
          colors = c("Egg"="#FFD301","Nymph"="#9BA4B5","female"="#EB455F","male"="#1982c4"),
          symbol= ~adult_sisters,
          symbols= c("yes"="circle", "no"= "circle-open"),
          text=~sample,
          hoverinfo= "text") %>%
  add_markers() %>%
  layout(title = 'F2 offspring of F1 male (1/1) x female (1/1)',
        scene = list(
        xaxis = list(title = '0/0', nticks = 3, range = c(0,1)),
        yaxis = list(title = '0/1', nticks = 3, range = c(0,1)),
        zaxis = list(title = '1/1', nticks = 3, range = c(0,1))))

# 2d
df <- left_join(samples, observed, by=c("sample","gt")) %>%  
    replace(is.na(.),0) %>% # replace "NA" with zero value when gt doesn't exists 
    group_by(sample) %>%
    mutate(total = as.numeric(sum(obs))) %>%
    mutate(freq=round(obs/total, digit=2)) %>%
  left_join(meta, by="sample")

p = ggplot(df, aes(x = sex, y = freq, colour = gt, shape = adult_sisters,
              text = paste("sample:", sample,
                          ", n:", total))) + 
  geom_boxplot(outlier.shape = NA, coef=0 ) +
  geom_jitter(width=0.1, size=2) +
  scale_color_manual(values=c("#ffbf00", "#66b032","#1982c4")) +
  theme_classic() +
scale_shape_manual(values=c(1,19)) +
  ggtitle('F2 offspring of F1 male (0/0) x female (0/0)') +
    facet_wrap( ~gt)  +  guides(color = "none") 
ggplotly(p) %>% 
  layout(legend = list(orientation = "h", x = 0, y = -0.1))
```

## Crossing homozygotic male, with heterozygotic female
### (5) F1 male (0/0) x female (0/1)
```{r}
# Prepare table with observed and expected counts, per family:
# define a list to put all the data frames in
obs <- list()

for (fam in family) {
 
obs[[fam]] <- table %>%
  dplyr::select(starts_with(fam)) %>%
        dplyr::filter_at(vars(matches("_fnd")), all_vars(. == "0/0")) %>% # force F0 female to be homo, like her son
dplyr::filter_at(vars(matches("_son")), all_vars(. == "0/0")) %>%
  dplyr::filter_at(vars(matches("_dat")), all_vars(. == "0/1")) %>% 
  dplyr::select(contains("grn")) %>%
  tidyr::pivot_longer(everything())  %>% 
  dplyr::rename(sample = name, gt = value) %>%
  dplyr::count(sample, gt, .drop = FALSE) %>% 
  dplyr::filter(gt %in% c("0/0", "1/1", "0/1")) %>%
  mutate(n = as.numeric(n)) %>%
  group_by(sample) %>%
#  mutate(total = as.numeric(sum(n))) %>%
  dplyr::rename(obs = n)
}

# bind all families together, to a final data frame containing all observed counts
observed <- do.call("rbind", obs) %>% mutate(sample = as.character(sample))
# replace "NA" with zero value when gt doesn't exists
samples <- data.frame(sample = rep(unique(observed$sample), each =3), gt = rep(c("0/0", "0/1", "1/1"))) # add all possible genotypes

# 3d
data <- left_join(samples, observed, by=c("sample","gt")) %>%  
    replace(is.na(.),0) %>% # replace "NA" with zero value when gt doesn't exists 
    group_by(sample) %>%
    mutate(total = as.numeric(sum(obs))) %>%
    mutate(freq=obs/total) %>% 
    pivot_wider(names_from="gt", values_from=freq, id_cols=c(sample,total)) %>%
    left_join(meta, by="sample") %>%
    rename(homo_ref="0/0") %>%
      rename(hetero="0/1") %>%
    rename(homo_alt="1/1")

head(data)

data %>% #dplyr::filter(sex == "male") %>%
#  filter(!sample =="502_503d_grn") %>%
#    filter(!sample =="43_44c_grn") %>%
  plot_ly(x= ~homo_ref, y =~hetero, z = ~homo_alt,
          color = ~sex, 
          colors = c("Egg"="#FFD301","Nymph"="#9BA4B5","female"="#EB455F","male"="#1982c4"),
          symbol= ~adult_sisters,
          symbols= c("with adult sister"="circle", "no adult sister"= "circle-open"),
          text=~sample,
          hoverinfo= "text") %>%
  add_markers() %>%
  layout(title = 'F2 offspring of F1 male (0/0) x female (0/1)',
        scene = list(
        xaxis = list(title = '0/0', nticks = 3, range = c(0,1)),
        yaxis = list(title = '0/1', nticks = 3, range = c(0,1)),
        zaxis = list(title = '1/1', nticks = 3, range = c(0,1))))

# 2d
df <- left_join(samples, observed, by=c("sample","gt")) %>%  
    replace(is.na(.),0) %>% # replace "NA" with zero value when gt doesn't exists 
    group_by(sample) %>%
    mutate(total = as.numeric(sum(obs))) %>%
    mutate(freq=round(obs/total, digit=2)) %>%
  left_join(meta, by="sample")

p = ggplot(df, aes(x = sex, y = freq, colour = gt, shape = adult_sisters,
              text = paste("sample:", sample,
                          ", n:", total))) + 
  geom_boxplot(outlier.shape = NA, coef=0 ) +
  geom_jitter(width=0.1, size=2) +
  scale_color_manual(values=c("#ffbf00", "#66b032","#1982c4")) +
  theme_classic() +
scale_shape_manual(values=c(1,19)) +
  ggtitle('F2 offspring of F1 male (0/0) x female (0/1)') +
    facet_wrap( ~gt)  +  guides(color = "none") 

#play plot:
p <- df %>%  filter(total>10) %>%
  ggplot(aes(x = sex, y = freq, colour = gt, shape = adult_sisters,
              text = paste("sample:", sample,
                          ", n:", total))) + 
  geom_boxplot(outlier.shape = NA, coef=0 ) +
  geom_jitter() +
  scale_color_manual(values=c("#ffbf00", "#66b032","#1982c4")) +
  theme_classic() +
scale_shape_manual(values=c(1,19)) +
  ggtitle('F2 offspring of F1 male (0/0) x female (0/1)') +
    facet_wrap( ~gt+sex)  +  guides(color = "none") 

ggplotly(p) %>% 
  layout(legend = list(orientation = "h", x = 0, y = -0.1))
```

### (6) F1 male (1/1) x female (0/1)
```{r}
# Prepare table with observed and expected counts, per family:
# define a list to put all the data frames in
obs <- list()

for (fam in family) {
 
obs[[fam]] <- table %>%
  dplyr::select(starts_with(fam)) %>%
        dplyr::filter_at(vars(matches("_fnd")), all_vars(. == "1/1")) %>% # force F0 female to be homo, like her son
dplyr::filter_at(vars(matches("_son")), all_vars(. == "1/1")) %>%
  dplyr::filter_at(vars(matches("_dat")), all_vars(. == "0/1")) %>% 
  dplyr::select(contains("grn")) %>%
  tidyr::pivot_longer(everything())  %>% 
  dplyr::rename(sample = name, gt = value) %>%
  dplyr::count(sample, gt, .drop = FALSE) %>% 
  dplyr::filter(gt %in% c("0/0", "1/1", "0/1")) %>%
  mutate(n = as.numeric(n)) %>%
  group_by(sample) %>%
#  mutate(total = as.numeric(sum(n))) %>%
  dplyr::rename(obs = n)
}

# bind all families together, to a final data frame containing all observed counts
observed <- do.call("rbind", obs) %>% mutate(sample = as.character(sample))
# replace "NA" with zero value when gt doesn't exists
samples <- data.frame(sample = rep(unique(observed$sample), each =3), gt = rep(c("0/0", "0/1", "1/1"))) # add all possible genotypes

data <- left_join(samples, observed, by=c("sample","gt")) %>%  
    replace(is.na(.),0) %>% # replace "NA" with zero value when gt doesn't exists 
    group_by(sample) %>%
    mutate(total = as.numeric(sum(obs))) %>%
    mutate(freq=obs/total) %>% 
    pivot_wider(names_from="gt", values_from=freq, id_cols=c(sample,total)) %>%
    left_join(meta, by="sample") %>%
    rename(homo_ref="0/0") %>%
      rename(hetero="0/1") %>%
    rename(homo_alt="1/1")

head(data)

data %>% 
  plot_ly(x= ~homo_ref, y =~hetero, z = ~homo_alt,
          color = ~sex, 
          colors = c("Egg"="#FFD301","Nymph"="#9BA4B5","female"="#EB455F","male"="#1982c4"),
          symbol= ~adult_sisters,
          symbols= c("with adult sister"="circle", "no adult sister"= "circle-open"),
          text=~sample,
          hoverinfo= "text") %>%
  add_markers() %>%
  layout(title = 'F2 offspring of F1 male (1/1) x female (0/1)',
        scene = list(
        xaxis = list(title = '0/0', nticks = 3, range = c(0,1)),
        yaxis = list(title = '0/1', nticks = 3, range = c(0,1)),
        zaxis = list(title = '1/1', nticks = 3, range = c(0,1))))

# 2d
df <- left_join(samples, observed, by=c("sample","gt")) %>%  
    replace(is.na(.),0) %>% # replace "NA" with zero value when gt doesn't exists 
    group_by(sample) %>%
    mutate(total = as.numeric(sum(obs))) %>%
    mutate(freq=round(obs/total, digit=2)) %>%
  left_join(meta, by="sample")

# compute median by group
median <- df %>% group_by(gt,sex) %>% summarise(med_val=median(freq))
 
p = ggplot(df, aes(x = sex, y = freq, colour = gt, shape = adult_sisters,
              text = paste("sample:", sample,
                          ", n:", total))) + 
  geom_boxplot(outlier.shape = NA, coef=0 ) +
  geom_jitter(width=0.1, size=2) +
  scale_color_manual(values=c("#ffbf00", "#66b032","#1982c4")) +
  theme_classic() +
scale_shape_manual(values=c(1,19)) +
  ggtitle('F2 offspring of F1 male (1/1) x female (0/1)') +
    facet_wrap( ~gt)  +  guides(color = "none") 
ggplotly(p) %>% 
  layout(legend = list(orientation = "h", x = 0, y = -0.1))


```

## Crossing heterozygotic male, with homozygotic female
### (7) F1 male (0/1) x female (0/0)
```{r}
# Prepare table with observed and expected counts, per family:
# define a list to put all the data frames in
obs <- list()

for (fam in family) {
 
obs[[fam]] <- table %>%
  dplyr::select(starts_with(fam)) %>%
        dplyr::filter_at(vars(matches("_fnd")), all_vars(. == "0/1")) %>% # force F0 female to be homo, like her son
dplyr::filter_at(vars(matches("_son")), all_vars(. == "0/1")) %>%
  dplyr::filter_at(vars(matches("_dat")), all_vars(. == "0/0")) %>% 
  dplyr::select(contains("grn")) %>%
  tidyr::pivot_longer(everything())  %>% 
  dplyr::rename(sample = name, gt = value) %>%
  dplyr::count(sample, gt, .drop = FALSE) %>% 
  dplyr::filter(gt %in% c("0/0", "1/1", "0/1")) %>%
  mutate(n = as.numeric(n)) %>%
  group_by(sample) %>%
#  mutate(total = as.numeric(sum(n))) %>%
  dplyr::rename(obs = n)
}

# bind all families together, to a final data frame containing all observed counts
observed <- do.call("rbind", obs) %>% mutate(sample = as.character(sample))
# replace "NA" with zero value when gt doesn't exists
samples <- data.frame(sample = rep(unique(observed$sample), each =3), gt = rep(c("0/0", "0/1", "1/1"))) # add all possible genotypes

data <- left_join(samples, observed, by=c("sample","gt")) %>%  
    replace(is.na(.),0) %>% # replace "NA" with zero value when gt doesn't exists 
    group_by(sample) %>%
    mutate(total = as.numeric(sum(obs))) %>%
    mutate(freq=obs/total) %>% 
    pivot_wider(names_from="gt", values_from=freq, id_cols=c(sample,total)) %>%
    left_join(meta, by="sample") %>%
    rename(homo_ref="0/0") %>%
      rename(hetero="0/1") %>%
    rename(homo_alt="1/1")

head(data)

data %>% 
  plot_ly(x= ~homo_ref, y =~hetero, z = ~homo_alt,
          color = ~sex, 
          colors = c("Egg"="#FFD301","Nymph"="#9BA4B5","female"="#EB455F","male"="#1982c4"),
          symbol= ~adult_sisters,
          symbols= c("with adult sister"="circle", "no adult sister"= "circle-open"),
          text=~sample,
          hoverinfo= "text") %>%
  add_markers() %>%
  layout(title = 'F2 offspring of F1 male (0/1) x female (0/0)',
        scene = list(
        xaxis = list(title = '0/0', nticks = 3, range = c(0,1)),
        yaxis = list(title = '0/1', nticks = 3, range = c(0,1)),
        zaxis = list(title = '1/1', nticks = 3, range = c(0,1))))

# 2d
df <- left_join(samples, observed, by=c("sample","gt")) %>%  
    replace(is.na(.),0) %>% # replace "NA" with zero value when gt doesn't exists 
    group_by(sample) %>%
    mutate(total = as.numeric(sum(obs))) %>%
    mutate(freq=round(obs/total, digit=2)) %>%
  left_join(meta, by="sample")

p = ggplot(df, aes(x = sex, y = freq, colour = gt, shape = adult_sisters,
              text = paste("sample:", sample,
                          ", n:", total))) + 
  geom_boxplot(outlier.shape = NA, coef=0 ) +
  geom_jitter(width=0.1, size=2) +
  scale_color_manual(values=c("#ffbf00", "#66b032","#1982c4")) +
  theme_classic() +
scale_shape_manual(values=c(1,19)) +
  ggtitle('F2 offspring of F1 male (0/1) x female (0/0)') +
    facet_wrap( ~gt)  +  guides(color = "none") 
ggplotly(p) %>% 
  layout(legend = list(orientation = "h", x = 0, y = -0.1))
```

### (8) F1 male (0/1) x female (1/1)
```{r}
# Prepare table with observed and expected counts, per family:
# define a list to put all the data frames in
obs <- list()

for (fam in family) {
 
obs[[fam]] <- table %>%
  dplyr::select(starts_with(fam)) %>%
        dplyr::filter_at(vars(matches("_fnd")), all_vars(. == "0/1")) %>% # force F0 female to be homo, like her son
dplyr::filter_at(vars(matches("_son")), all_vars(. == "0/1")) %>%
  dplyr::filter_at(vars(matches("_dat")), all_vars(. == "1/1")) %>% 
  dplyr::select(contains("grn")) %>%
  tidyr::pivot_longer(everything())  %>% 
  dplyr::rename(sample = name, gt = value) %>%
  dplyr::count(sample, gt, .drop = FALSE) %>% 
  dplyr::filter(gt %in% c("0/0", "1/1", "0/1")) %>%
  mutate(n = as.numeric(n)) %>%
  group_by(sample) %>%
#  mutate(total = as.numeric(sum(n))) %>%
  dplyr::rename(obs = n)
}

# bind all families together, to a final data frame containing all observed counts
observed <- do.call("rbind", obs) %>% mutate(sample = as.character(sample))
# replace "NA" with zero value when gt doesn't exists
samples <- data.frame(sample = rep(unique(observed$sample), each =3), gt = rep(c("0/0", "0/1", "1/1"))) # add all possible genotypes

data <- left_join(samples, observed, by=c("sample","gt")) %>%  
    replace(is.na(.),0) %>% # replace "NA" with zero value when gt doesn't exists 
    group_by(sample) %>%
    mutate(total = as.numeric(sum(obs))) %>%
    mutate(freq=obs/total) %>% 
    pivot_wider(names_from="gt", values_from=freq, id_cols=c(sample,total)) %>%
    left_join(meta, by="sample") %>%
    rename(homo_ref="0/0") %>%
      rename(hetero="0/1") %>%
    rename(homo_alt="1/1")

head(data)

data %>% 
  plot_ly(x= ~homo_ref, y =~hetero, z = ~homo_alt,
          color = ~sex, 
          colors = c("Egg"="#FFD301","Nymph"="#9BA4B5","female"="#EB455F","male"="#1982c4"),
          symbol= ~adult_sisters,
          symbols= c("with adult sister"="circle", "no adult sister"= "circle-open"),
          text=~sample,
          hoverinfo= "text") %>%
  add_markers() %>%
  layout(title = 'F2 offspring of F1 male (0/1) x female (1/1)',
        scene = list(
        xaxis = list(title = '0/0', nticks = 3, range = c(0,1)),
        yaxis = list(title = '0/1', nticks = 3, range = c(0,1)),
        zaxis = list(title = '1/1', nticks = 3, range = c(0,1))))
# 2d
df <- left_join(samples, observed, by=c("sample","gt")) %>%  
    replace(is.na(.),0) %>% # replace "NA" with zero value when gt doesn't exists 
    group_by(sample) %>%
    mutate(total = as.numeric(sum(obs))) %>%
    mutate(freq=round(obs/total, digit=2)) %>%
  left_join(meta, by="sample")

p = ggplot(df, aes(x = sex, y = freq, colour = gt, shape = adult_sisters,
              text = paste("sample:", sample,
                          ", n:", total))) + 
  geom_boxplot(outlier.shape = NA, coef=0 ) +
  geom_jitter(width=0.1, size=2) +
  scale_color_manual(values=c("#ffbf00", "#66b032","#1982c4")) +
  theme_classic() +
scale_shape_manual(values=c(1,19)) +
  ggtitle('F2 offspring of F1 male (0/1) x female (1/1)') +
    facet_wrap( ~gt)  +  guides(color = "none") 
ggplotly(p) %>% 
  layout(legend = list(orientation = "h", x = 0, y = -0.1))
```

## Crossing heterozygotic male and female
### (9) F1 male (0/1) x female (0/1)
```{r}
# Prepare table with observed and expected counts, per family:
# define a list to put all the data frames in
obs <- list()

for (fam in family) {
 
obs[[fam]] <- table %>%
  dplyr::select(starts_with(fam)) %>%
        dplyr::filter_at(vars(matches("_fnd")), all_vars(. == "0/1")) %>% # force F0 female to be homo, like her son
dplyr::filter_at(vars(matches("_son")), all_vars(. == "0/1")) %>%
  dplyr::filter_at(vars(matches("_dat")), all_vars(. == "0/1")) %>% 
  dplyr::select(contains("grn")) %>%
  tidyr::pivot_longer(everything())  %>% 
  dplyr::rename(sample = name, gt = value) %>%
  dplyr::count(sample, gt, .drop = FALSE) %>% 
  dplyr::filter(gt %in% c("0/0", "1/1", "0/1")) %>%
  mutate(n = as.numeric(n)) %>%
  group_by(sample) %>%
#  mutate(total = as.numeric(sum(n))) %>%
  dplyr::rename(obs = n)
}

# bind all families together, to a final data frame containing all observed counts
observed <- do.call("rbind", obs) %>% mutate(sample = as.character(sample))
# replace "NA" with zero value when gt doesn't exists
samples <- data.frame(sample = rep(unique(observed$sample), each =3), gt = rep(c("0/0", "0/1", "1/1"))) # add all possible genotypes

data <- left_join(samples, observed, by=c("sample","gt")) %>%  
    replace(is.na(.),0) %>% # replace "NA" with zero value when gt doesn't exists 
    group_by(sample) %>%
    mutate(total = as.numeric(sum(obs))) %>%
    mutate(freq=obs/total) %>% 
    pivot_wider(names_from="gt", values_from=freq, id_cols=c(sample,total)) %>%
    left_join(meta, by="sample") %>%
    rename(homo_ref="0/0") %>%
      rename(hetero="0/1") %>%
    rename(homo_alt="1/1")

head(data)

data %>% 
  plot_ly(x= ~homo_ref, y =~hetero, z = ~homo_alt,
          color = ~sex, 
          colors = c("Egg"="#FFD301","Nymph"="#9BA4B5","female"="#EB455F","male"="#1982c4"),
          symbol= ~adult_sisters,
          symbols= c("with adult sister"="circle", "no adult sister"= "circle-open"),
          text=~sample,
          hoverinfo= "text") %>%
  add_markers() %>%
  layout(title = 'F2 offspring of F1 male (0/1) x female (0/1)',
        scene = list(
        xaxis = list(title = '0/0', nticks = 3, range = c(0,1)),
        yaxis = list(title = '0/1', nticks = 3, range = c(0,1)),
        zaxis = list(title = '1/1', nticks = 3, range = c(0,1))))

# 2d
df <- left_join(samples, observed, by=c("sample","gt")) %>%  
    replace(is.na(.),0) %>% # replace "NA" with zero value when gt doesn't exists 
    group_by(sample) %>%
    mutate(total = as.numeric(sum(obs))) %>%
    mutate(freq=round(obs/total, digit=2)) %>%
  left_join(meta, by="sample")

p = ggplot(df, aes(x = sex, y = freq, colour = gt, shape = adult_sisters,
              text = paste("sample:", sample,
                          ", n:", total))) + 
  geom_boxplot(outlier.shape = NA, coef=0 ) +
  geom_jitter(width=0.1, size=2) +
  scale_color_manual(values=c("#ffbf00", "#66b032","#1982c4")) +
  theme_classic() +
scale_shape_manual(values=c(1,19)) +
  ggtitle('F2 offspring of F1 male (0/1) x female (0/1)') +
    facet_wrap( ~gt)  +  guides(color = "none") 
ggplotly(p) %>% 
  layout(legend = list(orientation = "h", x = 0, y = -0.1))
```

